---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='../Fisher-exact-test.md') })
output:
  html_document:
    highlight: tango
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path = "../img/Fisher-exact-test/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width = 100)
```

# Performing Fisher's exact test

In this tutorial we will be using data provided by [Freeman and Campbell from The University of Sheffield](https://www.sheffield.ac.uk/polopoly_fs/1.43998!/file/tutorial-9-fishers.pdf) that examined whether intra-muscular magnesium better treated chronic fatigue syndrome compared to a placebo treatment in 32 patients.  The results of the study are provided in the table below.

<div align="center">
<table style="width:60%">
  <tbody>
    <tr>
      <td></td>
      <td><b>Magnesium</b></td>
      <td><b>Placebo</b></td>
    </tr>
    <tr>
      <td><b>Felt better</b></td>
      <td><center>12</center></td>
      <td><center>3</center></td>
    </tr>
    <tr>
      <td><b>Did not feel better</b></td>
      <td><center>3</center></td>
      <td><center>14</center></td>
    </tr>
  </tbody>
</table>
</div>

For this example we will manually put the data into R rather than reading it from a file.  Using the `data.frame()` function we can build a data frame object where we input the data from top to bottom row-wise then left to right column-wise.  We can do so by first providing a name for the column, placing an equal sign, then using `c()` to record the data row-wise.  The `row.names` option can also be used to create a list of names for the rows in the data frame.

```{r data_input}
df <- data.frame(Magnesium = c(12, 3), Placebo = c(3, 14),
                row.names = c("Felt better", "Did not feel better"))

df
```

We can now run the test by wrapping the data frame in the `fisher.test()` function.  Like other functions we can change whether we want the test to be two- or one-sided through the `alternative` option, however because the magnesium treatment could have a better, worse, or no effect we should keep it to the default `two.sided`.

```{r fisher_test}
fisher.test(df)
```

From the results of the test we can see that with a p-value less than 0.05 (`p-value = 0.001033`) we can reject <i>H<sub>0</i></sub> to conclude that the rows and columns are not independent.  In the context of our data, we can conclude that patients given the magnesium treatment overall felt better than those who took the placebo.

### Full code block

```{r full_code, eval = FALSE}
# Create a data frame
df <- data.frame(Magnesium = c(12, 3),Placebo = c(3, 14),
                row.names = c("Felt better", "Did not feel better"))

df

# Run a two-sided Fisher's exact test
fisher.test(df)
```