---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='../multinomial-logistic-regression.md') })
output:
  html_document:
    highlight: tango
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="../img/multinomial-logistic-regression/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# Multinomial Logistic Regression in `R`

For this guide we will assess what factors influence an employee's choice on one of three health insurance plans offered by a company.  The [data set](https://github.com/tylerbg/DLC_stat_resources/tree/master/docs/R/dat/health_insurance.csv), which is a reduced version of the data set provided in the [Handbook of Regression Modeling in People Analytics](https://peopleanalytics-regression-book.org/index.html), has the following 4 factors:

* *plan* - the health insurance plan chosen (plan A, B, or C).
* *age* - the person's age when they chose their healthcare plan.
* *gender* - the gender of the person (Male or Female).
* *absent* - the number of absentee days from work the year prior to choosing their health insurance plan.

There are 50 observations evenly split between Male and Female employees.

To begin, we will load in the data using `read.csv()`, then take a quick look at the first few rows with `head()` and get a summary of the data set with `summary()`.

```{r load-data}
health <- read.csv("../../dat/health_insurance.csv")

head(health)
summary(health)
```

In the output from `summary()` we can see that the *gender* and *plan* variables were read as character variables, however we would want them to be factor where there are 2 levels in *gender*, Male and Female, and 3 levels in *plan*, A, B, and C. We can easily change these variables to factors using `factor()` to reassign each variable. We can then use `str()` to see the structure of the data, which will allow us to see the variable types and some other basic information on those variables.

```{r assign-factors}
health$gender <- factor(health$gender)
health$plan <- factor(health$plan)

str(health)
```

We see that *gender* and *plan* are now factors with 2 and 3 levels, respectively. We can also note that the *age* and *absent* variables were set as integers rather than numeric, which is fine for our purposes here.

Now we will fit our multinomial model where we have the health insurance plans, *plan*, as our dependent variable with *age*, *gender*, and *absent* as predictors. `R` does not have a base multinomial regression function so we will need to import a library to fit our model. Two popular packages for multinomial logistic regression are `nnet` and `VGAM`. In this example we will use `VGAM` which is more friendly with some of the functions we used to assess model results in our other guides.

After installing and loading the `VGAM` package we will use `vglm()`, a function to fit multiple types of Generalized Linear Models. In `vglm()` we put the model formula in the same format as other models where the dependent variable is on the left and predictor variables on the right of a `~`, with `.` as a shortcut to include all variables not included on the left side of the equation. We then assign our data set to `data` and assign `multinomial` to `family` to sepecify we aim to fit a multinomial regression model.

In addition, we will want to fit a null model (one withour our predictor variables and only an intercept) that we can use to test the fit of our model. We could either write a null model in `vglm()` or we can use `update()`, where we include the object that is our original model followed by the desired model formula, which for a null model we put a `1` on the right side of the `~` to indicate we only want an intercept. After both models are fit we will use `lrtest()`, a function provided with `VGAM`, to perform a likelihood ratio test to compare the two models.

```{r fit-mlr}
library(VGAM)

health_mlr <- vglm(plan ~ .,
                   data = health,
                   family = multinomial)

health_mlr0 <- update(health_mlr, . ~ 1)

lrtest(health_mlr0, health_mlr)
```

From the results of the likelihood ratio test we can see that the log-likelihood (`LogLik`) is higher in the model that includes our variables (second model in the output) compared to the null model. Importantly, this difference is significant with a very low p-value (`Pr(>Chisq)` = `1.228e-08`). We then have evidence that our model explains some of the variance compared to the null model and follow-up with the `summary()` function to assess some of the model's results.

```{r mlr-summary}
summary(health_mlr)
```

In the output we should first note the bottom line that says <code class="out">Reference group is level  3  of the response</code>, which corresponds to plan `C`. This line is telling us that the other levels, plans `A` (level 1) and `B` (level 2), are being compared to plan `C` in the `Coefficients` table. In that table we see that two coefficients are statistically significant, `age:1` and `genderMale:2`. As both have negative coefficients we can interpret these results as:

* As *age* increases the log-likelood of an employee choosing plan `A` over plan `C` decreases.
* `Male` employees are less likely to choose plan `B` over plan `C` compared to `Female` employees.

We can look at some plots of the data to see these differences more clearly.

```{r plots, echo = FALSE}
library(tidyverse)

p1 <- ggplot(health, aes(x = plan, y = age)) +
  stat_summary(geom = "pointrange",
               fun.data = mean_sdl,
               fun.args = list(mult = 1)) +
  stat_summary(geom = "errorbar",
               fun.data = mean_sdl,
               fun.args = list(mult = 1),
               width = 0.2) +
  geom_jitter(width = 0.1, height = 0,
              alpha = 0.2) +
  labs(x = "Plan",
       y = "Age",
       title = "Employees who pick one of three health\ninsurance plans based on their age.") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(health, aes(x = plan, fill = gender, color = gender)) +
  geom_bar(position = "dodge", fill = "white") +
  geom_bar(position = "dodge", alpha= 0.5) +
  labs(x = "Plan",
       y = "Count",
       title = "Number of employees who pick one of three health\ninsurance plans separated by gender.") +
  theme_bw() +
  scale_fill_discrete(name = "Gender") +
  scale_color_discrete(name = "Gender") +
  theme(plot.title = element_text(hjust = 0.5))

library(gridExtra)

grid.arrange(p1, p2, nrow = 1)
```



```{r contrasts-age}
library(multcomp)

K <- matrix(c(0, 0, -1, 1, 0, 0, 0, 0), 1)

summary(glht(health_mlr, linfct = K))
```

```{r contrasts-gender}

K <- matrix(c(0, 0, 0, 0, -1, 1, 0, 0), 1)

summary(glht(health_mlr, linfct = K))
```

```{r contrasts-absent}

K <- matrix(c(0, 0, 0, 0, 0, 0, -1, 1), 1)

summary(glht(health_mlr, linfct = K))
```