---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='../two-proportions.md') })
output:
  html_document:
    highlight: tango
editor_options:
  chunk_output_type: console
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path = "../img/two-proportions/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width = 100)
```

### Performing a test of equal proportions

For this R tutorial we will use the example from [Penn State's Introduction to Mathematical Statistics course](https://online.stat.psu.edu/stat415/lesson/9/9.4).  Assume that 800 American adults were polled on the question: "Should the federal tax on cigarettes be raised to pay for health care reform?".  Of those polled, 195 identified themselves as *smokers*, 41 of which responded 'yes', while 605 identified as *non-smokers* with 351 responding yes.  Plotting the results of the survey (below) we might suggest that there is a difference in how *smokers* and *non-smokers* responded overall, so we can use a proportion test to statistically support or reject this hypothesis.

```{r plot, echo = FALSE}
df <- data.frame(Status = c(rep("Non-Smoker", 2), rep("Smoker", 2)),
                 Answer = c(rep(c("Yes", "No"), 2)),
                 n = c(351, 254, 41, 154))

library(tidyverse)

ggplot(df, aes(x = Status, y = n, fill = Answer)) +
  geom_col() +
  scale_fill_brewer(palette = "Paired") +
  labs(title ="Reponses from 800 American adults to:\n\"Should the federal tax on cigarettes be raised to pay for health care reform?\"",
       x = "Smoking Status",
       y = "Count") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

In this tutorial we will manually input the data into R.  First, we will make two vectors, one with the total number of *non-smokers* and *smokers* polled (*count*) and the second with the number of yes responds for each group (*yes*) with the `c()` function.

```{r input_data}
n <- c(605, 195)
yes <- c(351, 41)
```

We can perform a proportion test using the `prop.test()` function which takes the vector of successes as the first argument and the total number of trials as the second argument.

```{r prop_test}
prop.test(yes, n)
```

The p-value for our test is extremely small and well below 0.05, so we can conclude that there is a statistically significant difference between the two proportions. Specifically, the proportion of smokers who answered 'yes' to the survey question is significantly lower than the proportion of non-smokers who answered 'yes'.

*Note that the test statistic for our test is slightly different than in the STAT 415 example. This is because `prop.test()` uses the Yates' continuity correction by default. By setting `correct = FALSE` within prop.test() the correction will not be made to result in the same test statistic that is calculated in the STAT 415 example.*

### Full code block

```{r full_code, eval=FALSE}
# Input the total number of non-smokers and somkers polled and the number of yes responses for each
n <- c(605, 195)
yes <- c(351, 41)

# Perform a proportion test
prop.test(yes, n)
```

