---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,  
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'data_cleaning.md')) })
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.align=TRUE)
options(width=125)
```

# Data Cleaning

###### Published: 6 April 2020

\newline

The first thing with any dataset, whether we collected the data ourselves or retrieve it from another source, is to thoroughly check the quality of and clean our data.  In fact, [preparing raw data is the most time consuming part of data analysis](https://visit.figure-eight.com/rs/416-ZBE-142/images/CrowdFlower_DataScienceReport_2016.pdf). It is important that we [follow certain steps to clean our data](https://sunscrapers.com/blog/why-is-clean-data-so-important-for-analytics-and-business-intelligence/) to reduce issues during data analysis, which include (but not always limited to):

- Identifying and correcting structural errors
- Handling missing values
- Removing duplicated observations
- Dealing with incorrect, mislabeled, and outlying data points
- Formatting consistent variable schemes, removing unwanted variables, and generating new variables

Here, we will go over some methods to identify and handle these issues in addition to some basics on:

- Loops and apply functions
- Pipes
- Pattern recognition and substitution

Before we attempt to clean our data we should first summarize and explore our dataset to get familiar with it and spot any possible errors or inconsistencies.

## Summarizing and Exploring Data

First, let us set up our working directory and load it.  Our data is stored as a *.csv* file in the "*data/*" folder, so we can access it using the `read.csv()` function.

```{r}
setwd("~/CDAR")
beer <- read.csv("data/beer_to_clean.csv")
```

To get familiar with our data we should check the size of our dataset.  If we use the `dim()` function we can see its dimensions.

```{r}
dim(beer)
```

The first number (2021) indicates the amount of rows, our total number of observations, and the second (10) the columns, the number of variables measured.  2021 is a lot of rows, much more than we can reasonably look through.  Instead, we can look at a few rows to get a general idea of what our data looks like.  To do so, we can use the `head()` and `tail()` functions to look at the first and last few rows, respectively.

```{r}
head(beer)
tail(beer, 3)
```

In the code above we used `head()` with just the name of our dataframe and we were given the first 6 rows of observations.  Instead, when we used the `tail()` function we included a 3 to tell `R` to give us the last 3 rows rather than the last 6 as default (this can be changed as much as you like to view more or less rows, but we will continue to use 3 for these functions just to reduce output space).

While both the `head()` and `tail()` function provide the column names, we can use a separate function `names()` (or alternatively `colnames()`) to return only those column names.

```{r}
names(beer)
```

For this dataset the column names are fairly straight-forward.  `brewery_name` is the name of the brewery, `review_appearance` is the score given by for how the reviewer viewed the appearance of the beer, and so on.  However, there is one column that does not make any sense, `X`.  If we look back up to our `head()` and `tail()` commands we can see that the `X` column increases incrementally from 1.  In fact, each number for this column is the name as their corresponding row number.  Maybe these are unique ID numbers, or maybe there was a formatting issue, but either way this column does not appear to be necessary.  Typically, once we have chosen that a column can be removed then we should remove it before continuing with other procedures so as not to negatively affect our data exploration.  So, let us go ahead and remove that column.

To do so, we can reassign our dataframe using a variety of methods.  One method is presented in the following code block.

```{r, eval=FALSE}
beer <- beer[, -1]
```

Here, we are telling `R` to reassign our dataframe using our old dataframe without the first column by using the -1.  

Alternatively, we could replace the brackets `[, -1]` with `[, 2:10]` to tell `R` to copy over each column starting from the 2nd to the 7th columns.

```{r}
beer <- beer[, 2:10]
names(beer)
```

However, make sure to only run either command once!  Running the first command a second time would delete the `brewery_name` column (which would not be the first column), while running the second would return an error.  Again using the `names()` function we can see that we have removed `X` from our dataframe but kept the other 9 columns.  

Now, it would be good to look at a summary of our dataset, which we can do with the `summary()` function.

```{r}
summary(beer)
```

As we can see, the `summary()` function provides the cut-off values for the 1st, 2nd (Median), and 3rd quartiles in addition to the minimum, maximum, and average (mean) values for numerical data.  We can use the summary statistics to get an overall idea how our data is distributed as well as discover possible problems.  One issue to point out is the maximum values for `beer_abv` of 89, a very large (and likely impossible) ABV for a beer!  

Are we certain this is the only outlier in this variable?  If we look at the second maximum we may be able to answer that question.  We can use the `sort()` command to help us out here, setting the `decreasing` option to `TRUE` with a value of 2 in the brackets `[]` that follow.  This tells `R` to sort the data in decreasing order and select the second value (our second maximum).

```{r}
sort(beer$beer_abv, decreasing=TRUE)[2]
```

The 15 seems fairly normal for a high ABV beer, so it appears there is only one outlier here.

Alternatively, if we were uncertain how many outliers we might have we could choose a threshold to look above (or below in other cases) and count the amount of data points that are outside of that threshold.  For a high ABV beer, we could assume that a maximum of 30 is a reasonable threshold.  Then, we can employ a little logic to add up all of the values that are above 30 with the following code.

```{r}
sum(beer$beer_abv > 30, na.rm=TRUE)
```

Only one beer is above our threshold, so again it looks like we have only one outlier here. Note that we needed to set the `na.rm` option to `TRUE`.  We have missing values in this variable which we will cover in a moment.

*Note: You may sometimes see `TRUE` and `FALSE` in `R` simplified to `T` and `F`, however [this is not recommended](https://www.r-bloggers.com/r-tip-avoid-using-t-and-f-as-synonyms-for-true-and-false/)*

But what if we were unfamiliar with beer and did not know if 15 was a normal ABV or not?  Instead of a threshold we set we could use statistical methods to determine if we have an outlier, such as the [Grubb's test](https://www.statisticshowto.com/grubbs-test/) or the [generalized extreme studentized deviate test](https://www.itl.nist.gov/div898/handbook/eda/section3/eda35h3.htm), which we will cover another time.  We can also identify outliers visually by plotting our data in boxplots, scatterplots, histograms, *etc*.  These will be covered in Exploratory Data Analysis.  How we will deal with this outlier will be explored in the next section.

Before moving forward we should bring up the `attach()` function which allows `R` to access variables directly from their names (without having to first type `beer$`).  So, we could alternatively run the code above as:

```{r}
attach(beer)
sort(beer_abv, decreasing=TRUE)[2]
sum(beer_abv > 30, na.rm=TRUE)
```

Notice, we no longer need to include `beer$` before the variable name and we get the same output.  While that makes our typing a little bit easier, using `attach()` [is not advised for multiple reasons](https://stackoverflow.com/questions/10067680/why-is-it-not-advisable-to-use-attach-in-r-and-what-should-i-use-instead) so we will avoid using the function here and in future posts.

Looking through the other variables in our summary statistics we notice the ratings for each type of review are on a scale from 1 to 5 in increments of 0.5.  Are we certain that there are not any incorrect values here?  The minimum and maximum values look fine, and so do the quartiles.  The means are not divisible by 0.5, however that is not a concern since it is a measure of the average score.  What else can we do to make sure there are no obvious issues in this variable?

One approach is to use some simple math and logic.  If we know that the scale is in increments of 0.5, then if we divide each score by 0.5 then they should all be whole numbers, and thus will not have remainders in their division.  Therefore, we can divide all of the values by 0.5 then add up each time the remainder is not 0 as shown in the code below (in `R` we use `%%` to give the remainder of division).

```{r}
sum(beer$review_aroma %% 0.5 > 0)
```

Our code returned a 0, meaning that there were no remainders to summate, so it looks like the `review_aroma` variable is solid.  We can perform the same procedure with the other review variables to confirm that their data at least appears correct.  We could copy the same code above and replace `review_aroma` for each of the other review variables, or we could use `apply()` to run our function on each column of interest and simplify our code.

```{r}
apply(beer[,c(3,4,6,7)], 2, function(x) sum(x %% 0.5 > 0))
```

Here, we first identify which data we will be applying our function on, then the 2 indicates that the function will be on the columns, then we describe the function we want to apply.  As we can see, each of our review variables return 0 so everything seems okay here.

However, not all of our data is numerical to provide measures of quartiles, means, *etc*. that based some of our initial observations on.  For instance, `brewery_name` gives the total counts for each brewery, but how did `R` know what to do here?  We can call another function `str()` to look at the structure of the dataframe our data is in.

```{r}
str(beer)
```

From this output we have three different types of data in our dataframe: 

- Integer (int) which are whole numbers 
- Numeric (num) which is also known as Double
- Factor

At first look it appears that Factors are Strings (or known in `R` as Characters), however that is not the case.  Instead, Factors are categorized data which fall into a number of levels.  For example, if we observe `brewery_name` in our output above we see the variable is of the Factor type with 5 levels.  We can use the `levels()` command to look further.

```{r}
levels(beer$brewery_name)
```

Now, we see that each of the 5 levels of the `brewery_name` variable is the name of one of our 5 breweries.  We can also see there are some issues in the names, but we will take care of that in the next section.

If we go back to our summary statistics there is one more important observation we should make.  In the `beer_abv` variable we are given a value for "NA's" of 17 (remember we needed to use the `na.rm` option previously).  This is telling us that there are 17 values that are missing in this variable, or are "Not Available".  Could there be other missing values in our dataset?  We can use the following command to find that out.

```{r}
sum(is.na(beer))
```

Here, we are using the function `is.na()` which returns `TRUE` for every value of "NA" in the dataset.  We use the `sum()` command to add up each time `is.na()` returned `TRUE` to get a total number of values in the dataset with "NA".  As we can see, this number matches the number of "NA's" in the `beer_abv` column so it does not appear that there are missing values elsewhere.  We could also use the following command to see the total number of "NA" values in each column.

```{r}
colSums(is.na(beer))
```

The `colSums()` function works similar to the `sum()` function, however it calculates the sum for each column rather than for the dataset overall.  Again, we can see that there are 17 missing values in `beer_abv` but none in the other columns.  If we were interested in looking at rows we could similarly use the `rowSums()` function.

While we do not observe them in our dataset here there are [three other types of missing data in `R`](https://www.r-bloggers.com/r-null-values-null-na-nan-inf/): `NULL`, `NA`, and `Inf`.  While slightly different, [there are functions to handle these missing value types](https://www.r-bloggers.com/r-null-values-null-na-nan-inf/) similar to those available for `NA` values.

So now that we know that we have `NA` values what should we do with them?  There are a few options, and we will cover them in the next section as well.

For now, let's check on one more thing, is any of our data duplicated?  Since each row represents a unique observation, we should check whether there are any non-unique rows.  To do so, we can use the `duplicated()` function, which returns `TRUE` whenever it finds a row that matches a previous row, along with the `sum()` function to count each time there is a duplicated row, similar to how we counted `NA` values above.

```{r}
sum(duplicated(beer))
```

It looks like we do have duplicated rows, 21 of them to be exact.  However, our function above only tells us the total amount of duplicates but not where they are in the dataset.  One way we can find the location of duplicates is through the `anyDuplicated()` function.

```{r}
anyDuplicated(beer)
```

From our output above, the first row that matches one of the preceding rows is row 2001.  If we do some simple math and logic, we could conclude that the last 21 rows of our dataset are duplicates.  Or, we could use a different function, `which()`, to instead show each duplicated row.

```{r}
which(duplicated(beer))
```

How convenient!  We could easily remove those 21 rows similar to how we removed the `X` column above, but what if the duplicates were interleaved in the dataset?  Using our previous method we would have to indicate every specific row we wanted to remove (an easy way to make mistakes).  While we could again employ the `which()` command, that approach could get messy.  Instead, we have a few other options which will start us off in our next section, cleaning up our dataset.

\newline

## Data Cleaning Procedures

#### Removing duplicates

Before we remove any duplicated observations we first must consider if our data could have duplicates by design.  For instance, if we asked customers to rank our beer on a scale from 1 to 10 but we did not keep track of each individual's response with some unique identifier, then we would expect some customers to give the same rating by chance, particularly if we survey a lot of customers.  Conversely, if we asked customers to rank 15 different attributes it is less likely that any two reviewers would score each attribute the same.

Our dataset observes individual reviews of beers, and since we only have four ratings on a scale of 1 to 5 in 0.5 increments, it is plausible that two independent reviewers could give the same beer the same score.  However, we have another variable, `review_time`, which records the time the review was made as a unique ID for every review.  Therefore, the duplicates in our dataset are true duplicates due to some error in data processing rather than by happenstance. 

There are multiple ways we can deal with duplicated rows, [such as using the `!` operator or the `distinct()` function](https://www.datanovia.com/en/lessons/identify-and-remove-duplicate-data-in-r/), however for our purposes the `unique()` function will do the trick of only keeping non-duplicated rows.

```{r}
beer <- unique(beer)

sum(duplicated(beer))

dim(beer)
```

We then again use the `sum()` and `duplicated()` functions to show we no longer have duplicated rows and the `dim()` function to see that we have 2021-21 = 2000 rows and 9 columns, which we expect. So now can move ahead to clean up specific values.

\newline

#### Outliers & misrepresented data

In our summary statistics table above we noticed that there appears to be an outlying observation in the `beer_abv` variable.  We can call the `which.max` function to tell us which specific observation (row number) has the outlier, then use that number to index the observation as shown in the following code. 

```{r}
max.abv <- which.max(beer$beer_abv)
max.abv

beer[max.abv, ]
```

The "*Tettnager Trippel*" in row 121 contains our outlier.  To correct it, we can first start by looking to see if there are any other "*Tettnager Trippel*" in the dataset.

```{r}
beer[beer$beer_name == "Tettnager Trippel", ]
```

It looks like there is only one observation for this beer, so we have a few choices.  We could:

- Discard this observation (generally, this should be our last option).
- Replace the ABV with the average or median ABV from the same beer style, "*Tripel*".  
- Since `beer_abv` is a descriptor variable in our dataset it is likely that we could look up the true value.

Whichever method we choose, since this is one observation in 2000 it likely will not make a huge difference, however in cases with smaller sample sizes this choice could be very significant.  If we look up this brand of beer we [can see that the actual ABV is 8.9%](https://www.beeradvocate.com/beer/profile/345/67529/?ba=glid02), therefore we can assume that there was a data input error and the actual value was supposed to be 8.9 rather than 89.  We can then reassign this single value with the following code.

```{r}
beer$beer_abv[max.abv] <- 8.9

summary(beer$beer_abv)
```

And from the summary statistics for the `beer_abv` variable, the distributions of the beer ABVs appears to be more understandable.  However, we still have 17 `NA` values to deal with.

\newline

#### Missing values

One way we can choose to deal with missing values is to replace them with a measure of central tendency.  In our summary statistics table we saw that the median and mean values for `beer_abv` were pretty similar, 7.00 and 7.236, respectively.  So if we went with this method it probably will not make a difference between which we choose, so we will stick with the average value.

*(NOTE:  We are not evaluating the following code block because we will be using a more efficient method ahead)*

```{r, eval=FALSE}
mean.abv <- mean(beer$beer_abv, na.rm=TRUE)

beer[is.na(beer)==TRUE] <- mean.abv
```

In the first line we assign the mean for `beer_abv`.  Again note that we need to set the `na.rm` option to `TRUE` in the `mean()` function to tell `R` to ignore missing values in the calculation, otherwise when we print out `mean.abv` we will also receive a value of `NA`.  Many functions in `R` have methods for dealing with missing values, however they typically do so by ignoring (or removing) those missing values which is generally not the best choice in data analysis.  Next, we assign all of the `NA` values with the mean ABV.

While the code above works for our dataset above, what would happen if we had missing values in other columns, such as in the `beer_style` variable?  We would have assigned `NA` values with the mean ABV which would have been incorrect!  An alternative way to perform the code above that only works on our column of interest is as follows:

```{r, eval=FALSE}
mean.abv <- mean(beer$beer_abv, na.rm=TRUE)

beer$beer_abv[is.na(beer$beer_abv)==TRUE] <- mean.abv
```

Here, we use the `$` operator after `beer` to tell `R` we are only working with the `beer_abv` column, so it will not change values of "NA" in other columns.

Alternatively, we could try to make our replacement values a little bit more accurate by using prior information.  For instance, we could replace the `NA` values with the median ABV for the respective style of beer.  That is, for our first observation with a `NA` we want to replace it with the median ABV for "Hefewizen", then the second observation with the median ABV for "German Pilsner", and so forth.  This would (ideally) be a slightly more accurate correction since different beer styles can vary in their range of ABVs.  To do this, we could use a `for()` loop.

```{r}
beer.na <- which(is.na(beer$beer_abv))
style.list <- as.character(beer$beer_style[beer.na])
```

First, we get two vectors, one with the row numbers of each `NA` value and one with the styles of beer that correspond to each of those missing values.

```{r, eval=FALSE}
median.abv <- rep(0, length(style.list))
for(i in 1:length(style.list)){
  median.abv[i] <- median(beer$beer_abv[beer$beer_style == style.list[i]], na.rm=TRUE)
}

beer$beer_abv[beer.na] <- median.abv
```

Above, we make a list of *0's* with the length of our vector of `NA` values.  Then, we run a `for()` loop to generate a new vector of values for the median ABVs.  Then, we replace the missing values with the median ABVs.  If we wanted to further check that our missing values are replaced correctly, we could again use the `sum()` and `is.na()` or we could print out the rows that we replaced those values for.

While this method works, the `for()` loop above is clunky and [many suggest avoiding](https://www.r-bloggers.com/for-loops-and-how-to-avoid-them/) `for()` loops in `R` as [much as possible](https://rpubs.com/markpayne/132503).  So, alternatively we can use the `sapply()` function to calculate the median value for each item in our list we made above.

```{r}
median.abv <- sapply(style.list, function(x) median(beer$beer_abv[beer$beer_style == x], na.rm=TRUE))

beer$beer_abv[beer.na] <- median.abv

sum(is.na(beer))
```

First, we use the vector we made earlier that has the styles of beer we are interested in to apply our function to collect the median ABV for each style.  Then, we write our function to replace the `NA` values in the dataframe with the median ABVs and use the `sum()` and `is.na()` functions to show that we no longer have missing values.

Alternatively, we could remove all of our observations that have missing values.  In our dataset, that would be 17 deletions of 2000 observations, so only 0.85% of our data, which could be reasonable.  However, in many situations deleting whole observations would be a much more significant loss of data so it would not be a viable option versus the other methods listed above (or others).

\newline

#### Renaming, moving, & creating variables

If it is not yet clear, it will be if we continue much further that keeping the variable names as they currently are is somewhat tedious to code.  Do we really need to add "*review_*" for each review variable?  Or have "*beer_*" for each descriptive variable when we know we are looking at data about beer?  How about we make the variable names easier to work with while we code by changing the column names.

```{r}
colnames(beer) <- c("brewery", "time", "aroma", "appearance", "style", "palate", "taste", "name", "abv")

names(beer)
```

Much better!

*Note: If we had used the `attach()` function we would need to reapply it here.  Otherwise, `R` will still pull the original columns*

Looking back at our data we can see that there is an awkward order of the variables.  Two of our review variables, `aroma` and `appearance` are in the beginning of the dataframe while two others, `palate` and `taste` are near the end.  While this does not technically matter it will be easier when we do our analysis later (and make more sense) if they are consecutive.  One way we can reorder our data is shown in the following code block.

```{r}
beer <- beer[, c(2, 1, 8, 5, 9, 3:4, 6:7)]

head(beer, 3)
```

What we have done now is move our unique identifier (`time`) to the first column, descriptive variables to the following columns, and our measured data (the review responses) to the last columns.  The data does not have to be organized in this way, however this is a fairly typical order of variables you will see and will tend to make coding easier in the future.

Alternatively, we could take advantage of one of the most used and most helpful packages (really group of packages) called `tidyverse`.  Using the `select()` function we can order our columns by their names rather than by their index.  While this sometimes takes more typing it generally makes more reproducible and easier to read code.

```{r}
library(tidyverse)
beer <- select(beer, time, brewery, name, style, abv, aroma, appearance, palate, taste)
```

Now that we have the columns in the order we want, how about we put the rows in a preferred order.  The values for the `time` variable does not mean too much for us at the moment, but we could assume that they are in ascending order of when the review was taken.  Therefore, why don't we order our rows using that variable (which we are using as our unique identifier anyway!).

```{r}
beer <- beer[order(beer$time), ]

head(beer, 3)
```

Now that our rows are ordered by `time` you might notice that our rownames are no longer in order.  We can easily rename (or renumber) the rows as well.

```{r}
rownames(beer) <- seq(1:2000)

head(beer, 3)
tail(beer, 3)
```

The `seq()` function above generates a sequence from 1 to 2000 at increments of 1 for our new row names.  Now our row labels and `time` are both in ascending order.

Next, we might want to consider adding other variables that could be of interest.  We have scores for 4 different attributes, but what about how the reviewers view each beer overall?  Since we do not have an overall ranking we can make our own based on the available data.  In the following code, we create two new variables that summate the total scores and take the average of all scores.

```{r}
beer$total <- rowSums(beer[, 6:9])
beer$average <- beer$total / 4

head(beer, 3)
```

Notice that since we reordered our variables we could use `[, 6:9]` to indicate the review variables rather than `[, c(3,4,7,8)]` if we had kept the columns in their original order.

While we would have preferred to have surveyed each reviewer on their overall liking of the beers, at least we can now get some idea of how beers are scored overall and which attributes are most important in that score.  

\newline

#### Recoding mislabeled data

Another problem we will commonly run across is mislabeled data, either mislabeled when the data was originally recorded or possibly during some format conversion.  As we can see in our data some of our mislabeling probably fall in the later category.  In particular, the "o" with an umlaut (&ouml;) in the name Tr&ouml;egs is replaced with other characters.  We could try to fix this by adding back the &ouml;, however we could continue to run into the same issue down the road and it would be much easier to just change it to a regular "o" for data analysis.

Also in the `tidyverse` package is a function called `recode_factor()` which will help us rename our mislabeled factors.  While we do not necessarily need this package to do so, renaming factors in `R` is not as straightforward or easy as we would generally like (remember they are not Characters or Strings but a special class), so we will use `recode_factor()` to simplify our work and our code.

```{r, message=FALSE}
beer$brewery <- recode_factor(beer$brewery,
                                   "Tröegs Brewing Company"="Troegs Brewing Company",
                                   "Yuengling Brewery "="Yuengling Brewery",
                                   "Yards Brewing Co."="Yards Brewing Company")
levels(beer$brewery)
```

On top of changing the special characters that represent the "&ouml;" to an "o", we also removed the space after "Yuengling Brewery" and spelled out the full name for "Yards Brewing Company".  If we left the space at the end of our factor name it could cause issues down the line if we try to call that level by name, and so to save the headache we should just change it now.  Why spell out the full name for company and not just leave the abbreviation?  Just to keep things consistent since the other levels had company spelled out.  Alternatively, we could have abbreviated the other levels to "Co." to shorten the name with the same function.

Finally, we again use the `levels()` function to print out the name for each level and see that our changes were successfully made.

While we fixed Tr&ouml;egs in the `brewery` column, we still are not done yet!  Tr&ouml;egs likes to add their company name to their beer names as well, which we can see if we take a look at all of the beer names in the `name` variable.  So, we can also replace the special characters with an "o" in this column as well.  However, if we used our code above we would have to go through and change multiple beer names that include Tr&ouml;egs, which would be time consuming.  Instead, we can use [pattern matching](https://bookdown.org/rdpeng/rprogdatascience/regular-expressions.html) to make our life (and code) simpler.

```{r, eval=FALSE}
beer$name <- as.character(beer$name)
beer$name <- gsub("Tröegs", "Troegs", beer$name)
beer$name <- as.factor(beer$name)
```

We first use `sum()` and `grepl()` to add up each time "Tröegs" appears in the column of beer names.  Then, we change the variable to a Character, use `gsub()` to match and replace each instance of "Tröegs" with "Troegs", then reassign the variable as a factor.  Finally, we again add up each time "Tröegs" appears to see that we completely converted the misspelling.

If we use the above code, we first change the `name` variable from a Factor to a Character then use the `gsub()` function to replace "*Tröegs*" with "*Troegs*".  We then turn `name` back into a Factor.  Alternatively, we could use [pipes](https://bookdown.org/rdpeng/rprogdatascience/managing-data-frames-with-the-dplyr-package.html#section) (`%>%`) and [regular expressions](https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf) to make coding a bit easier for us.

```{r}
beer$name <- beer$name %>%
  as.character() %>%
  gsub(pattern="Tr..egs", replacement="Troegs") %>%
  as.factor()

sum(grepl("Tr..egs", as.character(beer$name))==TRUE)
```

With pipes we only need to assign `beer$name` once then evaluate through.  Instead of including the specific "*ö*" characters, we use "*..*" which tells `R` that any character can be in that place.  Finally, we use `sum()` and `grepl()` commands to add up each time "Tr..egs" is matched.  Since it returned 0, we know we replaced all of the "Tröegs".

Now, let us take one "final" look at our data again using the `summary()` function.

```{r}
summary(beer)
```

From the summary table, it looks like we ahve cleaned up a lot of the issues!  We fixed the spelling of Tr&ouml;egs, removed duplicated rows, and no longer have missing values.  However, one thing to always keep in mind is that the summary table does not show you everything!  For example, let us take a look at the `style` column using the `levels()` function.

```{r}
levels(beer$style)
```

Uh oh, we have a few more instances where special characters did not format correctly with both K&oulm;lsch and M&oulm;rzen.  We could go ahead and fix those as well, but would it really matter?  If we are specifically interested in those two types of beer, or if we plan on displaying those beer styles in some figure or table then yes, we probably should correct them as well.  However, what if we are not interested in beer styles?  Or at least not observing those two styles specifically?  Then maybe it is not worth our time fixing them and we can leave them be.  But since we are here to learn about data cleaning in `R`, maybe it is worth our time and we can use the code below, similar to how we worked the same problem above.  And this time, we will just use pipes and regular expressions to make things (ideally) easier on us.

```{r}
beer$style <- beer$style %>%
  as.character() %>%
  gsub(pattern="M..rzen", replacement="Marzen") %>%
  gsub(pattern="K..lsch", replacement="Kolsch") %>%
  as.factor()

levels(beer$style)[c(50,54)]
```

As you can see above, we used the pipe function to first change the column to the Character class, then correct Marzen followed by Kolsch, then returned the column back to a Factor.  By using the brackets `[]` around the `c()` function after the `levels()` function we can tell `R` to only return our indices of interest rather than reprinting the entire list of 72 levels (keep in mind that after changing the Kolsch and Marzen text their level index slightly changed to 50 and 54 from their original indices of 49 and 53).

But what about `name`?  Surely some of these beers (and there are!).  Then we could reuse the piped code above but replace `style` with `name` after the `$`.

Now that we have a high quality dataset (or at least good enough for now) we should save it as a new file.

```{r}
write.csv(beer, "data/beer_cleaned.csv", row.names=FALSE)
```

Here, we are setting the `row.names` function to `FALSE`.  Otherwise, when we reload this data into `R` in the future we will again get a column titled `X` that lists the number of each row.

Are we really finished with data cleaning?  Probably not.  We will see when we continue with Exploratory Data Analysis (EDA) we may find additional issues with the data that we need to correct.

\newline

### Other Information and Resources

Code was generated in R version 3.6.3 (2020-02-29) -- "Holding the Windsock"

Data adapted from: https://www.kaggle.com/rdoume/beerreviews

[Cleaning data](https://www.theanalysisfactor.com/preparing-data-analysis/)

[Basics on working with data frames in `R`](https://www.r-bloggers.com/15-easy-solutions-to-your-data-frame-problems-in-r/)

[Data frame management](https://bookdown.org/rdpeng/rprogdatascience/managing-data-frames-with-the-dplyr-package.html)

[Data transformation](https://r4ds.had.co.nz/transform.html)

[Using lapply](https://www.r-bloggers.com/the-lapply-command-101/)

[Loop functions in `R`](https://bookdown.org/rdpeng/rprogdatascience/loop-functions.html)

[Pipes in `R`](https://r4ds.had.co.nz/pipes.html)

[Ways to deal with missing data](https://measuringu.com/handle-missing-data/)

[Data cleaning in python with Pandas](https://www.dataoptimal.com/data-cleaning-with-python-2018/)

[Data cleaning in `R` with tidyverse](https://towardsdatascience.com/data-cleaning-with-r-and-the-tidyverse-detecting-missing-values-ea23c519bc62)