# ---
# knit: (function(input_file, encoding) {
#   out_dir <- 'docs';
#   rmarkdown::render(input_file,  
#     encoding=encoding,
#     output_file=file.path(dirname(input_file), out_dir, 'distribution_test.md')) })
# output: github_document
# ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.align=TRUE)
options(width=125)
```

# Distribution Tests

###### Published: 10 April 2020

We will use a 

- age: age in years
- sex: sex (1 = male; 0 = female)
- cp: chest pain type
    - Value 1: typical angina
    - Value 2: atypical angina
    - Value 3: non-anginal pain
    - Value 4: asymptomatic
- trestbps: resting blood pressure (in mm Hg on admission to the hospital)
- chol: serum cholestoral in mg/dl
- fbs: (fasting blood sugar > 120 mg/dl) (1 = true; 0 = false)
- thalach: maximum heart rate achieved
- num: diagnosis of heart disease (angiographic disease status)
    - Value 0: < 50% diameter narrowing
    - Value 1: > 50% diameter narrowing

```{r}
setwd("~/CDAR")
hd <- read.csv("data/hd_cleveland.csv")
```

Of course, we should first start with taking a look at our data.

```{r}
head(hd)
summary(hd)
str(hd)
```

Comparing the output from the `str()` function with the descriptions of the variables above, we can tell that some variables were recognized as integers when they really should be factors.  We should change that before moving ahead.

```{r}
hd$sex <- factor(hd$sex)
hd$cp <- factor(hd$cp)
hd$fbs <- factor(hd$fbs)
hd$num <- factor(hd$num)
```



[Skewness and Kurtosis](https://www.r-bloggers.com/measures-of-skewness-and-kurtosis/) and coefficent of variation

```{r}
library(fBasics)

skewness(hd$age)
kurtosis(hd$age)

mean(hd$age) - skewness(hd$age)*sd(hd$age)/6
median(hd$age)

mean(hd$age) - skewness(hd$age)*sd(hd$age)/2
Mode(hd$age)

abs(mean(hd$age) - median(hd$age))
sd(hd$age)/mean(hd$age)
```

```{r}
Mode <- function(x){
    ta = table(x)
    tam = max(ta)
    if (all(ta == tam))
         mod = NA
    else
         if(is.numeric(x))
    mod = as.numeric(names(ta)[ta == tam])
    else
         mod = names(ta)[ta == tam]
    return(mod)
}
```


```{r}
hd.norms <- hd[, c(1,4,5,7)]
sapply(hd.norms, skewness)
sapply(hd.norms, kurtosis)
sapply(hd.norms, function(x) abs(mean(x) - median(x)))
sapply(hd.norms, function(x) mean(x) - skewness(x)*sd(x)/6)
sapply(hd.norms, median)
```


Shapiro-Wilk Normality Test

```{r}
shapiro.test(hd$age)
shapiro.test(hd$chol)
shapiro.test(hd$trestbps)
shapiro.test(hd$thalach)
```

From the Shipiro-Wilk Test, each of the four variables had low p-values (p<0.01).  We would therefore conclude that the distributions are not normal.

Histograms in Base `R`"

```{r}
par(mfrow=c(2,2))
sapply(hd.norms, function(x) hist(x, breaks=30))
par(mfrow=c(1,1))
```

Histograms in `ggplot2`:

```{r}
library(tidyverse)
hd.norms %>%
  gather(key="var", value="value") %>%
  ggplot(aes(x=value, fill=var)) +
  geom_histogram(bins=30) +
  facet_wrap(~var, scales="free") +
  theme_bw()
```

Q-Q plots in Base `R`:

```{r}
par(mfrow=c(2,2))
sapply(hd.norms, function(x) c(qqnorm(x), qqline(x), grid()))
par(mfrow=c(1,1))
```

Q-Q plots in `ggplot2`:

```{r}
library(ggpubr)
hd.norms %>%
  gather(key="var", value="value") %>%
  ggqqplot(x="value") +
  facet_wrap(~var, scales="free") +
  theme_bw()
```


```{r}
sort(hd$chol, decreasing=TRUE)[1:10]
which.max

shapiro.test(hd$chol[c(1:152, 154:303)])
```

```{r}
library(tidyverse)
hd[,c(1,4,5,8)] %>%
  gather(key="var", value="value") %>%
  ggplot(aes(x=value, fill=var)) +
  geom_histogram(bins=50) +
  facet_wrap(~var, scales="free")

shapiro.test(hd$chol[hd$chol<350])

```

```{r}
shapiro.test(log(hd$trestbps))
```

[Anderson-Darling](https://www.statology.org/how-to-conduct-an-anderson-darling-test-in-r/)

```{r}
library(nortest)

ad.test(hd$age)
```




Kolmogorov-Smirnov test

```{r}
ks.test(hd$age, "pnorm", mean(hd$age), sd(hd$age))
```

```{r}
ks.test(hd$chol, "pnorm", mean(hd$chol), sd(hd$chol))
ks.test(hd.norms$chol, "pnorm", mean(hd.norms$chol), sd(hd.norms$chol))
```

```{r}
set.seed(1)
hd.norms$chol <- hd.norms$chol + round(rnorm(303, mean=0, sd=0.5), 2)

ks.test(hd.norms$chol, "pnorm", mean(hd.norms$chol), sd(hd.norms$chol))
```


Fisher Exact Test

```{r}
hd.factors <- hd[, c(2,3,6,8)]

sex.cp.table <- table(hd.factors$sex, hd.factors$cp)
sex.cp.table

fisher.test(sex.cp.table)
```



Chi-sq test

```{r}
chisq.test(sex.cp.table)
```

D'Agostino's K-squared test

```{r}

```


### Other Resources

Data from [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Heart+Disease)

[Distributional Tests](https://www.itrcweb.org/gsmc-1/Content/GW%20Stats/5%20Methods%20in%20indiv%20Topics/5%206%20Distributional%20Tests.htm)

[Skewness & Kurtosis - SPC for Excel](https://www.spcforexcel.com/knowledge/basic-statistics/are-skewness-and-kurtosis-useful-statistics#skewness)

[Kurtosis - Stack Exchange](https://stats.stackexchange.com/questions/126346/why-kurtosis-of-a-normal-distribution-is-3-instead-of-0)

[Skewness & Kurtosis - codeburst](https://codeburst.io/2-important-statistics-terms-you-need-to-know-in-data-science-skewness-and-kurtosis-388fef94eeaa)