---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='ANCOVA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path = "img/ANCOVA/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width = 100)
```

# Analysis of Covariance (ANCOVA)

&emsp;Sometimes when we are using an ANOVA to test whether there are differences among some number of independent groups for some dependent variable we may believe that there is another continuous confounding variable that could affect our results.  In these cases we can adjust or control for this confounding variable through an ANCOVA, where instead of assessing differences between group means we statistically test for differences in the means that have been adjusted for the covariate.  With the ANCOVA we can have more than one continuous covariate that we control for, and like other ANOVA models the number of categorical independent variables determines if the ANCOVA is "one-way", "two-way", etc.

&emsp;With the ANCOVA we have similar assumptions to the ANOVA:

* The **dependent variable and the covariate(s) are continuous** while the **independent variable is categorical**
* The observations are **independent** and **randomly sampled**
* There are no **outliers** or **points with high leverage**
* The errors (residuals) are **approximately normally distributed** and have **equal variances**

&emsp;The null and altnerative hypotheses for the ANCOVA are also similar to those for the ANOVA, however the means are adjusted for the covariate.

&emsp;<i>H<sub>0</sub></i>: There is no effect of the independent variable on the dependent variable when controlling for the covariate.
<center>&mu;&#775;<sub>1</sub> = &mu;&#775;<sub>2</sub> = ... = &mu;&#775;<sub>k</sub></center>
<br>
&emsp;<i>H<sub>A</sub></i>: At least one of the adjusted group means is different from the other adjusted group means.
<center>&mu;&#775;<sub>i</sub> &ne; &mu;&#775;<sub>j</sub> for some i and j</center>
<br>

### Fitting an ANCOVA model

&emsp;[The data set for this example](https://github.com/tylerbg/DLC_stat_resources/blob/master/docs/R/dat/Rye-ANCOVA-2008.csv), originally provided in [STAT 485 Topics in R Statistical Language](https://online.stat.psu.edu/stat485/lesson/welcome-stat-485), compares the biomass of rye plants in g/m<sup>2</sup> (*RyeDMg*) planted at two different dates (*Plant*) and then terminated after a different number of days following the beginning of the growing season (*Tdate*).  After loaing the data we should get some information on the variables to make sure they are coded correctly and there are no issues such as missing data.  There are 5 other variables in the data set that do not interest us, such as the repetition number and term, so we will assign the variables that we are interested in to an object and print the structure and summary statistics of those variables as follows.

```{r read_rye}
rye <- read.csv("dat/Rye-ANCOVA-2008.csv", stringsAsFactors = TRUE)

cols <- c("Plant", "RyeDMg", "Tdate")

str(rye[, cols])
summary(rye[, cols])
```

&emsp;We can see that our dependent variable, *RyeDMg* is numerical (continuous) while our explanatory variable, *Plant*, is a factor (categorical) and the covariate *Tdate* is also numerical.  Therefore, our data set is coded correctly to fit a one-way ANCOVA.  Additionally, the summary statistics do not show any obvious outliers, missing data, or mis-coded observations, so we can continue with planning our analysis.

&emsp;Plotting the data (below) we can see that rye planted in early autumn (*P1*) appear to have a higher biomass throughout the growing season compared to those planted in late autumn (*P2*).  The rate of increase for these two planting times also appear to be very similar as their slopes are nearly parallel, suggesting there is not an interaction.

```{r rye_plot, echo = FALSE}
library(tidyverse)

ggplot(rye, aes(x = Tdate, y = RyeDMg, color = Plant)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.55, #position = position_dodge(width = 0.35),
             size = 3) +
  theme_bw() +
  labs(x = "Days from the beginning of the growing season (April 15)",
       y = bquote("Rye biomass"~(g/m^2))) +
  scale_color_manual(values = c("steelblue3", "tomato3"))
```

&emsp;To fit an ANCOVA model we can use the `aov()` function with the dependent variable to the left and explanatory variables to the right of a `~` in the formula.  Then, we can use the `summary()` or `anova()` function to print summary statistics of the model.

```{r rye_aov}
rye.aov <- aov(RyeDMg ~ Tdate + Plant, data = rye)

summary(rye.aov)
```

&emsp;From the ANCOVA sum of squares table we can see that both of the independent variables are statistically significant predictors of rye biomass.

&emsp;Note that the default sum of squares calculated by the `summary()` and `anova()` functions are Type I, or the sequential sum of squares.  This is fine for the example here, however when we include interaction terms or have unbalanced designs we may wish to reorder the terms to best reflect the experiment and research question.  To calculate the Type II or Type III sum of squares we can call functions in other libraries, such as `Anova()` from the `car` package.

&emsp;Another thing to note is that the ANCOVA model is mathematically the same as a multiple linear regression (MLR) model, and therefore using the `aov()` function with `lm()` will result in the same sum of squares table.  The difference between the ANCOVA (and other ANOVA methods) and MLR approaches is whether the sum of squares or statistics on the variable coefficients best answer our research question.

### Full Code 

```{r full_code, eval = FALSE}
# Load data set, assign variables of interest to an object, and print the structure and summary
# statistics of those variables
rye <- read.csv("dat/Rye-ANCOVA-2008.csv", stringsAsFactors = TRUE)

cols <- c("Plant", "RyeDMg", "Tdate")

str(rye[, cols])
summary(rye[, cols])

# Fit an ANCOVA model with RyeDMg as the dependent and Tdate and Plant as explanatory variables then
# print the 1st sum of squares
rye.aov <- aov(RyeDMg ~ Tdate + Plant, data = rye)

summary(rye.aov)
```