---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='binary-logistic-regression.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path="img/binary-logistic-regression/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width = 100)
```

# Binary Logistic (Logit) Regression

&emsp;For this example we will use this [data set](https://github.com/tylerbg/DLC_stat_resources/tree/master/docs/R/dat/leukemia_remission.csv) originally provided in [STAT 501 Regression Methods](https://online.stat.psu.edu/stat501/lesson/15/15.1) that includes six possible factors that may predict Leukemia remission:

* *BLAST* - absolute number of blasts in the peripheral blood
* *CELL* - cellularity of the marrow clot section
* *INFIL* - percentage of absolute marrow leukemia cell infiltrate
* *LI* - percentage labeling index of the bone marrow leukemia cells
* *SMEAR* - smear differential percentage of blasts
* *TEMP* - highest temperature prior to start of treatment

&emsp;We can load the data into R with the `read.csv()` function.  As always, it is good to look at the structure of the data set which we can do with `str()`.

```{r read_data}
remiss <- read.csv("dat/leukemia_remission.csv")

str(remiss)
```

&emsp;Our response variable for leukemia remission, *REMISS*, is either a 0 for no remission or a 1 for remission, meaning that it is binary and suitable for logistic regression.  Our six predictor variables are all continuous and we can leave them as they are.  We will go ahead and accept that the observations are independent, there are no outliers, and we have a sufficiently large sample size.

&emsp;Now that we have our data set loaded we can simply fit a logistic regression model with the `glm()` function.  We set up our model formula with the response variable, *REMISS*, to the left and the predictor variables to the right of a *~*. Instead of writing out all of our predictor variables we can instead choose to use a "." which will add all of the rest of the variables in the data frame to the model formula.  Finally, after filling out the `data =` argument we need to set `family = binomial` to indicate that we want to fit a logistic regression model, otherwise `glm()` would default to fitting a linear model.

&emsp;Like any other model fit by the `glm()` function we can use the `summary()` command to print the results.

```{r logit_fit1}
remiss.fit <- glm(REMISS ~ ., data = remiss, family = binomial)

summary(remiss.fit)
```

&emsp;The p-values for the Wald's tests on the estimated coefficients are listed in the `Pr(>|z|)` column, from where we can see that none of the six variables are significant predictors for leukemia remission in this model.  Unlike with Minitab or other statistical software, the base `summary()` function in R does not provide variance inflation factors (VIFs).  To calculate the VIFs for our fit logit model we can use the `vif()` function from the `car` package.

```{r vif}
library(car)

vif(remiss.fit)
```

&emsp;VIFs greater than 10 (or as some suggest greater than 5) indicate a high degree of multicollinearity among those variables which can severely affect regression models.  There are different suggestions on how to deal with high VIFs, but in our case we will choose to fit a simpler model with only one variable that we hypothesize would have the biggest influence on the probability of Leukemia remission, *LI*.

&emsp;Instead of recoding a new model in `glm()`, we can use `update()` to refit the original model with only *LI* as a predictor variable.  To do so, we input the object that we assigned our original model to then write the formula `. ~ LI` to indicate we wish to keep the response variable and only include LI as a predictor variable.  We can then assign the model to a new object and use `summary()` to print the results.

```{r logit_fit2}
remiss.fit2 <- update(remiss.fit, . ~ LI)

summary(remiss.fit2)
```

&emsp;In the summary from our simpler model we see that when *LI* is statistically significant with a p-value below 0.05. Keep in mind that the interpretation of the coefficient is different from linear regression. Specifically for our example, we could conclude that for every 1 unit increase in the *LI* variable the log odds of *REMISS* being 1 increases by 2.897. This type of interpretation will not be very meaningful for many applications, but the important information to take away here is that *LI* is a significant predictor of *REMISS* for which they share a positive relationship.  Or alternatively, as the percentage labeling index of the bone marrow leukemia cells increases so does the probability of leukemia remission, which we can observe as the red line in the plot below.

```{r remiss_plot, echo = FALSE, fig.height = 5}
plot(remiss$LI, remiss$REMISS, pch = 19, col = "white",
     xlab = "Percentage labeling index of the bone marrow leukemia cells (LI)",
     ylab = "Probability of leukemia remission (REMISS)",
     main = "Line plot of the fitted binary logistic regression model")
grid()
points(remiss$LI, remiss$REMISS, pch = 19, col = rgb(0.31, 0.58, 0.8, alpha = 0.41))
LI <- remiss$LI
remiss.preds <- curve(predict(remiss.fit2, data.frame(LI = x), type="resp"), add = TRUE,
                      col = "tomato3", lwd = 2)
# segments(0, 0.5, 1.3, 0.5, lty = 2)
# segments(1.3, 0, 1.3, 0.5, lty = 2)
```

### Full code block

```{r full_code, eval = FALSE}
# Load the data and display the structure of the resulting data frame
remiss <- read.csv("dat/leukemia_remission.csv")

str(remiss)

# Fit a binary logistic regression model including all predictor variables and print a summary of
# the results
remiss.fit <- glm(REMISS ~ ., data = remiss, family = binomial)

summary(remiss.fit)

# Load the car package to calculate VIFs for the full model
library(car)

vif(remiss.fit)

# Fit a new logistic regression model with only the LI variable and print a summary of the results
remiss.fit2 <- update(remiss.fit, . ~ LI)

summary(remiss.fit2)
```



