---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='chi-sq-test.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="img/chi-sq-test/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# The Chi-square (&Chi;<sup>2</sup>) Test

&emsp;When our data includes two categorical variables, we can determine if there is a statistically significant relationship (or alternatively dependence) between them with the Chi-square (&Chi;<sup>2</sup>) test. WIth the &Chi;<sup>2</sup> test, we compare the observed counts (our data) with the expected counts if the data came from the same distributions.

&emsp;For this tutorial we will use data from an example offered in [STAT 500 Applied Statistics](https://online.stat.psu.edu/stat500/lesson/8) where participants were asked to give their party affiliation (*Democrat* or *Republican*) and their opinion on a tax reform bill (*Favor*, *Indifferent*, *Opposed*).

| | Favor | Indifferent | Opposed |
| :--- | :---: | :---: | :---: |
| Democrat | 138 | 83 | 64 |
| Republican | 64 | 67 | 84 |

&emsp;The researcher wants to know whether a relationship exists between party affiliation and opinion so will test the following hypotheses:

<center><i>H<sub>0</sub></i>: No relationship exists between party affiliation and opinion on the tax reform bill</center>
<center><i>H<sub>A</sub></i>: There is a significant relationship between party affiliation and opinion on the tax reform bill</center>

### Enter the data

&emsp;There are many ways to enter data into R and mostly centers around preference. Two common ways to enter data into base R are the `matrix()` and `data.frame()` functions. Note that while they will produce the same table, they are of different data structures which can be important. However, either will work here.

### Data types

#### Matrix

```{r}
opinion <- matrix(data = c(138, 64, 83, 67, 64, 84),
                  nrow = 2, ncol = 3,
                  dimnames = list(c("Democrat", "Republican"),
                                  c("Favor", "Indifferent", "Oppose")))
```

#### data.frame()

```{r eval = FALSE}
opinion <- data.frame(Favor = c(138, 64),
                      Indifferent = c(83, 67),
                      Opposed = c(64, 84),
                      row.names = c("Democrat", "Republican"))


```

### Run the Chi-square test

&emsp;To run the Chi-square test we simply use the `chisq.test()` function on our dataset, either as a data frame or matrix.

```{r}
chisq.test(opinion)
```



&emsp;The output of the `chisq.test()` function gives us the test statistic (`X-squared = 22.152`), the degrees of freedom (`df = 2`), and the p-value associated with the test statistics (`p-value = 1.548e-05`). Importantly for answering our original question, the p-value is much less than 0 so that we can conclude that an association does exist between party affiliation and a person's opinion on the tax reform bill.


<table>
<tr><th>Observed </th><th>Expected</th></tr>
<tr><td>

| | Favor | Indifferent | Opposed | Total |
| :--- | :---: | :---: | :---: | :---: |
| Democrat | 138 | 83 | 64 | 285 |
| Republican | 64 | 67 | 84 | 215 |
| Total | 202 | 150 | 148 | 500 |

</td><td>

| | Favor | Indifferent | Opposed |
| :--- | :---: | :---: | :---: |
| Democrat | 115.14 | 85.5 | 84.36 |
| Republican | 86.86 | 64.5 | 63.64 |

</td></tr> </table>



<!-- Tab links -->
<div class="tab">
  <button class="tablinks" onclick="openCity(event, 'London')">London</button>
  <button class="tablinks" onclick="openCity(event, 'Paris')">Paris</button>
  <button class="tablinks" onclick="openCity(event, 'Tokyo')">Tokyo</button>
</div>

<!-- Tab content -->
<div id="London" class="tabcontent">
  <h3>London</h3>
  <p>London is the capital city of England.</p>
</div>

<div id="Paris" class="tabcontent">
  <h3>Paris</h3>
  <p>Paris is the capital of France.</p>
</div>

<div id="Tokyo" class="tabcontent">
  <h3>Tokyo</h3>
  <p>Tokyo is the capital of Japan.</p>
</div>