---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='Fisher-exact-test.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path = "img/Fisher-exact-test/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width = 100)
```

# Fisher's Exact Test

### Fisher's exact test hypotheses

&emsp;**Null hypothesis:** The proportions of the two variables are the equal.
<center><i>H<sub>0</sub>: p<sub>1</sub> = p<sub>2</sub></i></center>

&emsp;**Alternative hypothesis:** The proportions of the two variables are not equal.
<center><i>H<sub>A</sub>: p<sub>1</sub> &ne; p<sub>2</sub></i></center>
<br>

### [Fisher's exact test assumptions](https://online.stat.psu.edu/stat504/lesson/3/3.3)

- The total counts are **fixed and not random**
- **Small sample size**
- The samples and observations are **independent** of each other
- The observations are **mutually exclusive** and can be classified into only one category
- The **expected counts is < 5 but > 1** for at least 20% of the observations

### Entering the data in R

```{r}
x <- data.frame(tea_first = c(3, 1),
                milk_first = c(1, 3),
                row.names = c("tea", "milk"))
```

### Running the test & interpreting the results

&emsp;We can now run the test using the `fisher.test()` function, which like the proportions test is very simple and does not require us to calculate the proportions before hand and only uses one line of code.

```{r}
fisher.test(x, alternative = "greater")
```

&emsp;From the results of our test, we can see that our p-value of 0.2429 is greater than 0.05, indicating that there is not sufficient evidence to reject our null hypothesis and we can conclude that by taste alone the lady is not able to tell whether tea or milk was poured first.

### Full code block

```{r eval = FALSE}
# Enter the data
x <- data.frame(tea_first = c(3, 1),
                milk_first = c(1, 3),
                row.names = c("tea", "milk"))

# Run a Fisher's exact test
fisher.test(x, alternative = "greater")
```





```{r}
df <- data.frame(Female = c(9, 26), Male = c(21, 35),
                 row.names = c("Candidate A", "Candidate B"))

df
```

```{r}
fisher.test(df)
```

