---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='Kruskal-Wallis.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path = "img/Kruskal-Wallis/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width = 100)
```

# Kruskal-Wallis rank sum test

&emsp;For this example we will be using the `chickwts` data set supplied in base R.  The data set includes chick weights measured in grams (*weight*) for 71 chicks after being fed one of six supplemented feeds (*feed*) and are presented in the figure below.

```{r plot_data, echo = FALSE}
library(tidyverse)

ggplot(chickwts, aes(x = feed, y = weight, color = feed)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, fill = "grey97") +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  labs(title = "Chick weights by feed type after six weeks of feeding",
       x = "Feed supplement", y = "Chick weight (g)") +
  theme_bw() +
  theme(legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5))
```

&emsp;From the plot above we might hypothesize that there are differences in the chick weights after being fed some of these supplemented feeds for 6 weeks.  In particular, we might suggest that the horsebean feed leads to lower chick weights than the other feeds.

&emsp;Before we start fitting a statistical model to the data we should get an idea of what it looks like.  Using the `summary()` function we can print summary statistics of each variable.

```{r data_summary}
summary(chickwts)
```

&emsp;From the print out we can see that the *weight* variable has a wide range, from 108 grams to 423 grams.  For the *feed* variable we see that we have 6 feed types that are unequal but relatively close in size.  We do not see any miscoded data, missing values, obvious ouliers or skewness, or anything obvious that would cause concern.  For this example we will also accept that the assumptions for the Kruskal-Wallis test are valid for this data set, so we can continue with the Kruskal-Wallis test.

&emsp;To do so, we employ the `kruskal.test()` function that has the formula with the syntax "continuous response variable ~ categorical independent variable" that we will assign to an object (`chickwts.kw`).  Then to get the results of the test we can simply print the object as follows:

```{r kruskal_test}
chickwts.kw <- kruskal.test(weight ~ feed, chickwts)

chickwts.kw
```

&emsp;The `kruskal.test()` function provides the test statistic (`chi-squared = 37.343`), degrees of freedom (`df = 5`), and p-value (`p-value = 5.113e-07`), from which we can reject the null hypotheses at a statstical threshold of &alpha; = 0.05 to conclude that there is at least one group median that is different from the others.  However, similar to the one-way ANOVA the Kruskal-Wallis rank sum test does not tell us which median(s) is different from the others.  We will need to use a post-hoc test to make pairwise comparisons to determine which medians are statistically different from one another.  Instead of Tukey's honestly significant difference test for the parametric one-way ANOVA, we can use Dunn's test to make these non-parametric pairwise comparisons.

&emsp;In base R we do not have a function to perform the Dunn's test, so we will need to rely on the `FSA` package from which we can use the `dunnTest()` function with the same formula we used for the `kruskal.test()` function to assign the results to an object and then print the results directly from that object.

```{r dunn_test}
library(FSA)

chickwts.dunn <- dunnTest(weight ~ feed, chickwts)

chickwts.dunn
```

&emsp;Because we had 6 groups for pairwise comparisons we have $n*(n-1) / 2 = 6*(6-1) / 2 = 15$ total comparisons.  As you can imagine, as the number of groups increases the number of pairwise comparisons nearly exponentially increases.  The `dunnTest()` function already handles multiple test error correction for us (which we can choose to change the method of correction), however it can still be tedious to determine which p-values are statistical significant.  To make things easier we can use the `ifelse()` function to create a new column that indicates which comparisons are below our statistical threshold of &alpha; = 0.05 as follows:

```{r sig_values}
chickwts.dunn$res$sig <- ifelse(chickwts.dunn$res$P.adj > 0.05, "", "*")

chickwts.dunn
```

&emsp;Now it is easy to see that 5 of the 15 pairwise comparisons are statistically significant and we can make the following conclusions:

* Casein supplemented feed provided higher chick weights than the horsebean or linseed supplement
* The meatmeal supplemented feed also resulted in higher chick weights than horsebean
* Sunflower supplemented food results in higher chick weight compared to horsebean and linseed

### Full code block

```{r full_code, eval = FALSE}
# Print summary statistics of the chickwts data set
summary(chickwts)

# Perform a Kruskal-Wallis rank sum test and print the results
chickwts.kw <- kruskal.test(weight ~ feed, chickwts)

chickwts.kw

# Load the FSA library to run and print the results from the Dunn's post-hoc pairwise test
library(FSA)

chickwts.dunn <- dunnTest(weight ~ feed, chickwts)

chickwts.dunn

# Assign asterisks in a new column to note which pairwise comparisons are statistically significant
chickwts.dunn$res$sig <- ifelse(chickwts.dunn$res$P.adj > 0.05, "", "*")

chickwts.dunn
```

