---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='t-tests.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE,
                      fig.path="img/t-tests/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# T-tests

&emsp;When we are interested in testing whether one mean is equal to some value or the equality of two means we can rely on the t-test. While there are different forms of the t-test, they all are based on the t-distribution for a given mean and variance.

### T-test for one mean

&emsp;The simplest form of the t-test is when we have one sample for which we want to determine if the group is significantly different from some predetermined value. For instance, say we wanted to determine if the average student at Penn State had a normal blood cholesterol value (<200 mg/dL). At the beginning of the semester we recruited 10 students to donate their blood to our study and measured their cholesterol levels.

```{r one_mean}
PSU_student <- c(170, 230, 175, 145, 149, 165, 164, 150, 167, 151, 178, 183)

t.test(PSU_student, mu = 200, alternative = "less")
```

```{r one_mean_plot, echo = FALSE, fig.width = 4, out.extra='style="float:left; padding:10px"'}
# plot(density(PSU_student),
#      main = "Density plot of blood cholesterol levels for PSU students",
#      xlab = "Blood cholesterol",
#      col = "steelblue3",
#      lwd = 2)
# abline(v = 200, lty = 3, lwd = 2, col = "tomato")

stripchart(PSU_student, vertical = TRUE, pch = 16, cex = 2,
           col = rgb(.275, .51, .706, alpha = 0.55),
           main = "Student blood cholesterol levels\nat the beginning of the semester",
           ylab = "Blood cholesterol level (mg/dL)")
abline(h = 200, lty = 3, lwd = 2, col = "tomato")
abline(h = mean(PSU_student), lty = 2, lwd = 2, col = "steelblue3")
```

&emsp;We use the `t.test()` function on our dataset, `PSU_student`, and set the mean we are test to 200 with the `mu` option. Additionally, because we are only interested in if the students have blood cholesterol levels that are below that threshold, we set the `alternative` option to `"less"`, indicating that our alternative hypothesis for this test is that the true mean is less than the mean that we defined in the `mu` option as 200.  If instead we did not hypothesize that blood cholesterol would be more or less than 200 but not 200, we would set `alternative` to `"two.sided"` (the default value).

&emsp;From our results we see that the p-value for our t-test (p = 3.18 x 10<sup>-4</sup>) is well below our threshold of &alpha; = 0.05. Therefore, we can accept our alternative hypothesis that the mean blood cholesterol level for PSU students is less than 200 mg/dL.  The `t.test()` function also returns the mean of our sample, 168.9 mg/dL.

### Two-sample t-test

&emsp;Now let us say that we want to compare the average PSU student's blood cholesterol with the faculty at Penn State.

```{r faculty_data}
PSU_faculty <- c(189, 165, 191, 193, 177, 200, 179, 189, 202, 235, 178, 226)
```

```{r two_mean_plot, echo = FALSE, fig.width = 5}
PSU_stu_fac <- list("student" = PSU_student, "faculty" = PSU_faculty)
stripchart(PSU_stu_fac, vertical = TRUE, pch = 16, cex = 2,
           col = rgb(c(.275, .706), c(.51, .275), c(.706, .51), alpha = c(0.55, 0.55)),
           main = "Student and faculty blood cholesterol levels\nat the beginning of the semester",
           ylab = "Blood cholesterol level (mg/dL)",
           at = c(1.25, 1.75))
abline(h = mean(PSU_faculty), lty = 2, lwd = 2, col = "#b44682")
abline(h = mean(PSU_student), lty = 2, lwd = 2, col = "steelblue3")
```

&emsp;We can perform a two-sample t-test in R using the same `t.test()` function, but this time instead of defining `mu` we include the two data sets to be compared.

```{r two_means}
t.test(PSU_student, PSU_faculty)
```

&emsp;Because we are testing two groups we do not need to set a value for `mu` (the default is 0, meaning that we are testing if the difference of blood cholesterol values between students and faculty is or is not 0). Additionally, because we are interested in testing if there is a difference between the two groups in any direction we do not need to define the `alternative` option which is set to `"two.sided"` as default. Note that the `t.test()` function assumes the variances are unequal (`var.equal = FALSE` as default).

&emsp;From the output we can see that there is a statistically significant difference in the blood cholesterol levels between Penn State students and faculty (p = 0.01862). Additionally, the output provides a confidence interval from which we can conclude with 95% certainty that the mean difference between the two groups is somewhere between -45.99 and -4.81. Higher (or lower) ranges for the confidence interval can be set through the `conf.level` option.

### Paired t-test

&emsp;Now let us say that we collected the same group of students at the end of the semester during finals week to see if their blood cholesterol levels might have changed due to the stress of the semester.  We can compare the means from the beginning of the semester and those at finals week, but now we will be performing a paired t-test because the individuals we are measuring from are the same.  Note that it is important that we put the data in the same order!

```{r student2_data}
PSU_student2 <- c(179, 224, 175, 140, 148, 181, 179, 142, 180, 169, 170, 189)
```

```{r paired_plot, echo = FALSE, fig.width = 5}
PSU_stu_stu2 <- list("Beginning" = PSU_student, "Finals" = PSU_student2)
stripchart(PSU_stu_stu2, vertical = TRUE, pch = 16, cex = 2,
           col = rgb(c(.275, .706), c(.51, .471), c(.706, .275), alpha = c(0.55, 0.55)),
           main = "Student blood cholesterol levels\nat the beginning of the semester\nand during finals week",
           ylab = "Blood cholesterol level (mg/dL)",
           at = c(1.25, 1.75))
abline(h = mean(PSU_student2), lty = 2, lwd = 2, col = "#b47846")
abline(h = mean(PSU_student), lty = 2, lwd = 2, col = "steelblue3")
```

&emsp;We can again use the same `t.test()` function setting the `paired` option to `TRUE`.  Because we do not make any hypotheses that the blood cholesterol levels are higher or lower at the end of the semester, only different, we keep the `alternative` option to its default, `"two.sided"`.

```{r paired_test}
t.test(PSU_student, PSU_student2, paired = TRUE)
```

&emsp;The results from our test suggest that there is not a difference in student blood cholesterol levels from the beginning of the semester to finals week.  First, the p-value is above our threshold of 0.05.  In addition, the 95% confidence intervals around the mean of the differences (-4.08) includes 0 between the lower and upper limits.

### Full Code Block

```{r eval = FALSE}
# Input student blood cholesterol data from the beginning of the semester
PSU_student <- c(170, 230, 175, 145, 149, 165, 164, 150, 167, 151, 178, 183)

# Perform a one-mean t-test to determine if the average blood pressure is less than 200
t.test(PSU_student, mu = 200, alternative = "less")

# Add the blood cholesterol data from faculty
PSU_faculty <- c(189, 165, 191, 193, 177, 200, 179, 189, 202, 235, 178, 226)

# Compare student and faculty cholesterol levels with a two-sample t-test
t.test(PSU_student, PSU_faculty)

# Add cholesterol levels from the same students during finals week
PSU_student2 <- c(179, 224, 175, 140, 148, 181, 179, 142, 180, 169, 170, 189)

# Use a paired t-test to compare student cholesterol levels at the beginning and end of
# the semester
t.test(PSU_student, PSU_student2, paired = TRUE)
```

