---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='two-way-ANOVA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.path = "img/two-way-ANOVA/",
                      fig.width = 6, fig.height = 4,
                      fig.show = 'hold', fig.align = 'center')
options(width=100)
```

# Two-way Analysis of Variance (ANOVA<sup>2</sup>)

&emsp;For data sets with two factors the ANOVA<sup>2</sup> can be a powerful statistical tool when the data meets the following assumptions:

* The data are continuous,
* randomly sampled from the population,
* approximately normally distributed,
* and have equal variances between the groups.

&emsp;For this example we will use a data set from [STAT 414 Introduction to Mathematical Statistics](https://online.stat.psu.edu/stat415/lesson/14/14.1) where a physiologist wanted to determine if smoking history affected how a subject responds to different types of stress tests. The time to maximum oxygen uptake (VO<sub>2max</sub>) was measured from subjects belonging to three different categories of smoking history (*Nonsmoker*, *Moderate*, and *Heavy*) while performing three different stress tests (*Bicycle*, *Step Test*, and *Treadmill*).

```{r two-way-plot, echo = FALSE}
library(tidyverse)
stress <- read_csv("dat/smoking_stress.csv")

ggplot(stress, aes(x = Test, y = Time, color = `Smoking History`)) +
  geom_point(position = position_dodge(0.65),
             alpha = 0.5, size = 5) +
  theme_classic() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("steelblue", "goldenrod", "tomato3")) +
  xlab(NULL) +
  ylab("Time (min)") +
  ggtitle("Effect of stress tests and smoking history\n on the time to maximum oxygen uptake")
```

<br>

&emsp;The physiologist wants to test the following questions and their respective hypotheses:

1. Does smoking history affect the time to VO<sub>2max</sub>?
<center><i>H<sub>0</sub></i>:<i> &alpha;<sub>Nonsmoker</sub> = &alpha;<sub>Moderate</sub> = &alpha;<sub>Heavy</sub> = 0</i></center>
<center><i>H<sub>A</sub></i>: at least one <i>&alpha;<sub>i</sub> &ne; 0</i></center>
2. Are there differences between the types of stress tests and the time to reach VO<sub>2max</sub>?
<center><i>H<sub>0</sub></i>:<i> &beta;<sub>Bicycle</sub> = &beta;<sub>Step Test</sub> = &beta;<sub>Treadmill</sub> = 0</i></center>
<center><i>H<sub>A</sub></i>: at least one <i>&beta;<sub>j</sub> &ne; 0</i></center>
3. Is there an interaction between the three types of stress tests and smoking history?
<center><i>H<sub>0</sub></i>: <i>&alpha;&beta;<sub>ij</sub> = 0</i></center>
<center><i>H<sub>A</sub></i>: <i>&alpha;&beta;<sub>ij</sub> &ne; 0</i></center>

### Fit a two-way ANOVA

&emsp;First, we should read the data into R using the `read.csv()` command putting the pathway where our data set is stored in either double (`""`) or single (`''`) quotes.  We can also set `stringsAsFactors = TRUE` to force R to read the character variables as factors so that later coding will be more convenient.

```{r}
stress <- read.csv("dat/smoking_stress.csv", stringsAsFactors = TRUE)
```

&emsp;To fit a two-way ANOVA in R we can use the `aov()` command.  Since we will be calling other functions we should assign the results of the ANOVA<sup>2</sup> as an object, which we store here as `stress.aov`.

```{r}
stress.aov <- aov(Time ~ Smoking.History * Test, data = stress)
```

&emsp;As with coding other models in R, the response variable (*Time*) is coded to the left of a `~` while predictor variables (*Smoking History* and *Test*) are to the right. We separate the predictors with an asterisk `*` to indicate that we want to include both variables and an interaction term between them. If we wanted to include each term separately we can write the model as follows, which is equivalent to using the `*`.

```{r eval = FALSE}
stress.aov <- aov(Time ~ Smoking.History + Test + Smoking.History:Test, data = stress)
```

### Interpreting the results

&emsp;Using the `summary()` command we can print a table of results from our fitted ANOVA<sup>2</sup> model.

```{r}
summary(stress.aov)
```

&emsp;Notably, both *Smoking History* and *Test* are statistically significant with p-values well below our statistical threshold of &alpha; = 0.05.  However, the interaction between these two variables is not statistically significant.  We can then answer the physiologist's third question by concluding that there is not a significant interaction between the types of stress tests and a subject's smoking history.

&emsp;Sometimes you will see that researchers refit a model by dropping the interaction term when it turns up insignificant.  However, since the physiologist originally hypothesized an interaction between our two variables, we should leave it in the model and continue with pairwise comparisons on *Smoking History* and *Test*. 

### Post-hoc tests

&emsp;Since the interaction term is insignificant we should not follow up on that part of the ANOVA model with post-hoc tests.  So, to leave out a lot of unnecessary output we can first assign the results from the Tukey post-hoc test using `TukeyHSD()` to an object then use `$` to only print the pairwise comparisons from *Smoking History* and *Test*.

```{r}
stress.Tukey <- TukeyHSD(stress.aov)

stress.Tukey$Smoking.History
stress.Tukey$Test



```

&emsp;From the output, we can see that each pairwise comparison is statistically significant except for the comparison between *Nonsmoker* and *Moderate*. Therefore, we can answer the physiologist's original two questions as follows:

1. Heavy smokers have a significantly lower time to VO<sub>2max</sub> compared to moderate and nonsmokers, however there is not a statistically significant difference between moderate and nonsmokers.
2. There are statistically significant differences in the time to reach VO<sub>2max</sub> among the three types of stress tests, with the bicycle being the lowest, then treadmill, and the step test with the highest time.

### Full code block

```{r eval = FALSE}
# Read the data into R
stress <- read.csv("dat/smoking_stress.csv")

# Fit a two-way ANOVA with interaction effects then print the results
stress.aov <- aov(Time ~ Smoking.History * Test, data = stress)

summary(stress.aov)

# Perform a Tukey post-hoc test on the significant factors for pairwise comparisons
TukeyHSD(stress.aov)

stress.Tukey$Smoking.History
stress.Tukey$Test
```

