---
knit: (function(input_file, encoding) {
  out_dir <- 'docs/R';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='two-way-ANOVA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="img/two-way-ANOVA/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# Two-way ANOVA

### Load the data

&nbsp;&nbsp;&nbsp;&nbsp;For this example we will use a dataset from [STAT 414 Introduction to Mathematical Statistics](https://online.stat.psu.edu/stat415/lesson/14/14.1) where a physiologist wanted to determine if smoking history affected how a subject responds to different types of stress tests. The time to maximum oxygen uptake (VO<sub>2max</sub>) was measured from subjects belonging to three different categories of smoking history (*Nonsmoker*, *Moderate*, and *Heavy*) while performing three different stress tests (*Bicycle*, *Step Test*, and *Treadmill*). The physiologist wants to test the following questions and their respective hypoetheses:

1. Does smoking history affect the time to VO<sub>2max</sub>?
<center><i>H<sub>0</sub></i>:<i> &alpha;<sub>Nonsmoker</sub> = &alpha;<sub>Moderate</sub> = &alpha;<sub>Heavy</sub> = 0</i></center>
<center><i>H<sub>A</sub></i>: at least one <i>&alpha;<sub>i</sub> &ne; 0</i></center>
2. Are there differences between the types of stress tests and the time to reach VO<sub>2max</sub>?
<center><i>H<sub>0</sub></i>:<i> &beta;<sub>Bicycle</sub> = &beta;<sub>Step Test</sub> = &beta;<sub>Treadmill</sub> = 0</i></center>
<center><i>H<sub>A</sub></i>: at least one <i>&beta;<sub>j</sub> &ne; 0</i></center>
3. Is there an interaction between the three types of stress tests and smoking history?
<center><i>H<sub>0</sub></i>: <i>&alpha;&beta;<sub>ij</sub> = 0</i></center>
<center><i>H<sub>A</sub></i>: <i>&alpha;&beta;<sub>ij</sub> &ne; 0</i></center>

&nbsp;&nbsp;&nbsp;&nbsp;First, we should read the data into R using the `read.csv()` command and setting `stringsAsFactors = TRUE` so that R will know to read the character variables as factors.

```{r}
stress <- read.csv("dat/smoking_stress.csv", stringsAsFactors = TRUE)
```

### Fit the two-way ANOVA

&nbsp;&nbsp;&nbsp;&nbsp;To fit a two-way ANOVA in R we can use the `aov()` command as follows:

```{r}
stress.aov <- aov(Time ~ Smoking.History * Test, data = stress)
```

&nbsp;&nbsp;&nbsp;&nbsp;As with coding other models in R, our response variable *Time* will be on the left of a `~` while our predictor variables, *Smoking History* and *Test*, are to the right. We separate the predictors with an asterisk `*` to indicate that we want to include both variables and an interaction term between them. If we wanted to include each term separately we can write the model as follows, which is equivalent to using the `*`.

```{r eval = FALSE}
stress.aov <- aov(Time ~ Smoking.History + Test + Smoking.History:Test, data = stress)
```

### Interpreting the results

&nbsp;&nbsp;&nbsp;&nbsp;Using the `summary()` command we can output the results of our fitted model.

```{r}
summary(stress.aov)
```

&nbsp;&nbsp;&nbsp;&nbsp;Notably, both *Smoking History* and *Test* are statistically significant in the model with p-values well below 0.05. However, the interaction between these two variables is not statistically significant, so for the physiologist's third question we can conclude that there is not a significant interaction between the types of stress tests and a subject's smoking history.

&nbsp;&nbsp;&nbsp;&nbsp;Sometimes you will see that researchers refit a model by dropping the interaction term when it turns up insignificant. However, since the physiologist originally hypothesized an interaction between our two variables, we should leave it in the model and continue with pairwise comparisons on *Smoking History* and *Test*. 

&nbsp;&nbsp;&nbsp;&nbsp;Since the interaction term is insignificant we will not be following up on that part of the ANOVA model with post-hoc tests. So, to leave out a lot of unnecessary output we can first assign the results from the Tukey post-hoc test using `TukeyHSD()` to a variable, then using `$` to see just the pairwise comparisons from *Smoking History* and *Test*.

```{r}
stress.Tukey <- TukeyHSD(stress.aov)

stress.Tukey$Smoking.History
stress.Tukey$Test
```

&nbsp;&nbsp;&nbsp;&nbsp;From the output, we can see that each pairwise comparison is statisticall significant except for the comparison between *Nonsmoker* and *Moderate*. Therefore, we can answer the physiologist's original two questions as follows:

1. Heavy smokers have a significantly lower time to VO<sub>2max</sub> compared to moderate and nonsmokers, however there is not a statistically significant difference between moderate and nonsmokers.
2. There are statistically significant differences in the time to reach VO<sub>2max</sub> among the three types of stress tests, with the bicycle being the lowest, then treadmill, and the step test with the highest time.

### Full code block

```{r eval = FALSE}
# Read the data into R
stress <- read.csv("dat/smoking_stress.csv")

# Fit a two-way ANOVA with interaction effects then print the results
stress.aov <- aov(Time ~ Smoking.History * Test, data = stress)

summary(stress.aov)

# Perform a Tukey post-hoc test on the model for pairwise comparisons
TukeyHSD(stress.aov)

stress.Tukey$Smoking.History
stress.Tukey$Test
```

