---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='CO2_LR.md') })
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="imgs/CO2_LR/",
                      fig.width=6, fig.height=6,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# Analyzing `CO2` with linear regression

Linear regression models have the standard equation:

$${y_i = {\beta}_0 + {\beta}_1 x_{i,1} + {\beta}_2 x_{i,2} + ... + {\beta}_n x_{i,n} + {\epsilon}_i}$$

Where:
- $y_i$ is the response variable with $i$ observations,
- $x_i$ are predictor variables for an $n$ number of predictors and $i$ observations,
- ${\beta}$ are coefficients for $n$ predictor variables and the intercept, $\beta_0$,
- ${\epsilon}_i$ is the error term for $i$ observations.

In our case, fitting the response, `uptake`, with two categorical (`Type` and `Treatment`) and one continuous variable (`conc`) will have the equation:

$$y_{i, \textit{uptake}} = {\beta}_0 + {\beta}_\textit{Treatment} x_{i,\textit{Treatment}} + {\beta}_\textit{Type} x_{i,\textit{Type}} + {\beta}_\textit{conc} x_{i,\textit{conc}} + {\epsilon}_i$$

When we wish to include categorical data (factors) in a linear regression model we need to [recode the variables](https://stats.idre.ucla.edu/spss/faq/coding-systems-for-categorical-variables-in-regression-analysis/) to numerical values so that they may be entered into the regression equation and solved for.  Conveniently, `R` automatically dummy codes factors when they are included in a regresson model, so we can go ahead to fit and analyze the model using the following code.

```{r}
# Fit a linear model with uptake as the response variable
CO2.lm <- lm(uptake~Type+Treatment+conc, data=CO2)

# Print summary statistics for the linear model
summary(CO2.lm)
```

Taking a quick look at the summary statistics we can see that the linear model has a statistically significant fit overall (`p < 2.2e-16`) with a reasonably high R<sup>2</sup> (0.6839) and [adjusted R<sup>2</sup>](https://online.stat.psu.edu/stat501/lesson/10/10.3) (0.6721).  From the p-values for our estimated coefficients (`Pr(>|t|)`), we can see that each of the estimated coefficients are significant predictors of CO<sub>2</sub> uptake.  Therefore, we can conclude that our three predictors variables, `conc`, `Treatment`, and `Type`, are significant predictors of `uptake` and we can insert the coefficients into the linear regression equation to:

$${y_{i, \textit{uptake}} = 29.260 - 12.660 x_{i,\textit{Type}} - 6.860 x_{i,\textit{Treatment}}  + 0.018 x_{i,\textit{conc}}}$$

From these results we can conclude that <b> when `Type` and `Treatment` remain unchanged, for every one unit increase in `conc`, `uptake` increases by 0.018 umol/m<sup>2</sup></b>.  While it is easy to make conclusions for continuous variables like `conc`, the categorical variables `Type` and `Treatment` take slightly more work.  First, we need to see how `R` coded the categorical variables which we can do by calling `$xlevels` from the `lm()` object we just created.

```{r}
CO2.lm$xlevels
```

From the output we can see that for the Type and Treatment variables `Quebec` and `nonchilled` are first followed by `Mississippi` and `chilled`, respectively.  Knowing that R uses dummy coding for factors in regression we can surmise that the first levels are coded as 0 and the second as 1 for the two variables, which we can confirm when we use `summary()` to see the results of our regression model.

Notice that the Type and Treatment variables are concatenated with `Mississippi` and `chilled`, respectively.  When Type is `Quebec`, the dummy code is 0 and therefor the value will be 0 in the regression equation ($\beta_3 * x_{i,\textit{Type}} = -12.659 * 0 = 0$).  Conversely, when Type is `Mississippi` which is dummy coded to 1 the value for that variable will equal the estimated coefficient ($\beta_3 * x_{i,\textit{Type}}=-12.659 * 1 = -12.659$).  If we happened to have a third level in Type, the value for that factor would be double the estimated coefficient, triple for a fourth level, and so forth.

Therefore, we can conclude that when `Type` is `Mississippi`, `uptake` is reduced by 12.660 umol/m<sup>2</sup> compared to when `Type` is `Quebec` when the other variables remain constant.  Similarly, `chilled` decreases `uptake` by 6.860 umol/m<sup>2</sup> compared to `nonchilled` while `Type` and `conc` remain constant.

However, if we take a look back at our plots we may be convinced that there is an [interaction](https://online.stat.psu.edu/stat501/lesson/8/8.6) in the data, particularly between our two categorical variables `Type` and `Treatment`.  Information may be lost by not including this interaction in our model, so we should fit a linear regression model that includes an interaction term and assess its significance (or lack thereof).  We can include interaction terms in `R` formulas two ways:

- Use `*` between variables to include their stadard terms and interaction terms (`Type*Treatment)`).
- Use `+` for standard variable terms and `:` for interaction terms (`Type+Treatment+Type:Treatment`).

The `*` is the simplest way to include variables with interactions and becomes very useful when we have a lot of terms to include where the `+` and `:` method would take a lot of typing and increase our chance for errors.

Interaction terms are the product of a coefficient and the variables invovled in the interaction, which would make our regression equation for modeling CO<sub>2</sub> uptake:

$$y_{i, \textit{uptake}} = {\beta}_0 + {\beta}_\textit{Type} x_{i,\textit{Type}} + {\beta}_\textit{Treatment} x_{i,\textit{Treatment}} + {\beta}_\textit{conc} x_{i,\textit{conc}} + \beta_\textit{Type*Treatment}x_{i, \textit{Type}}x_{i, \textit{Treatment}} + {\epsilon}_i$$

We code this in `R` and print a summary of the results with:

```{r}
CO2.lm2 <- lm(uptake~Type*Treatment+conc, data=CO2)

summary(CO2.lm2)
```

The new model that includes the interaction term is still statistically significant (`p < 2.2e-16`), however we achieved a slightly higher R<sup>2</sup> and adjusted R<sup>2</sup> to indicate that we have a slightly better fitting model when including the interaction term.  We should also note that the estimates for the coefficients and their test statistics have changed.  In particular, `Treatment` is no longer statistically significant although the interaction term between `Type` and `Treatment` is.  This indicates that there is an interaction between these two factors, and even though `Treatment` is no longer statistically significant conventionally we should keep it in our model since it has a significant interaction with another variable.

We can also compare our first model without interactions against our new model that includes the `Type:Treatment` interaction to see if there is a statistically significant improvement in the model with `anova()`.

```{r}
anova(CO2.lm, CO2.lm2)
```

The second model that includes the interaction term has a significantly reduced RSS (residual sum of squares) as indicated by the p-value (`Pr(>F)`).  We can therefore conclude that the linear model is improved by adding the `Type:Treatment` interaction term.  If we did not see a statistically significant improvement in the model we should instead stick with the simpler model.

Our regression equation can now be written as:

$$y_{i, \textit{uptake}} = 27.621 - 9.381 x_{i,\textit{Type}} - 3.581 x_{i,\textit{Treatment}} + 0.018 x_{i,\textit{conc}} - 6.557 x_{i, \textit{Type}}x_{i, \textit{Treatment}} + {\epsilon}_i$$

Now that we know that there is an interaction between at least two of our variables our interpretation for the linear regression model that predicts `uptake` changes.  Specifically, we cannot claim that when our `Treatment` variable changes from `Quebec` to `Mississippi` the expected change in `uptake` decreases by 9.381 umol/m<sup>2</sup></b> (the coefficient for `Treatment` in our model with the interaction) when all other variables are constant because when the `Treatment` variable changes so does its respective interaction with `Type`.  Instead, the expected value for `uptake` changes by -9.381 - 6.557 * `Treatment` umol/m<sup>2</sup></b>. 

Like any other statistical model there is some degree of error in our estimation, so we may be interested in how confident we are in our estimates.  Using `confint()` we can print the 95% confidence intervals for the estimates of our regression coefficients.

```{r}
confint(CO2.lm2)
```

When interpreting the confidence intervals we can state that **with 95% confidence a unit increase in `conc` increases `uptake` by between 0.013 and 0.022 umol/m<sup>2</sup> when all other variables remain constant**.

We may also be interested into how well our model fits the data to determine if the linear regression appropriately fits or if a different approach may be necessary.  There are many methods of analyzing the fit of a model, one common and simple approach is to [assess the plots of the residual errors](https://data.library.virginia.edu/diagnostic-plots/).  In `R`, using the `plot()` function on an `lm()` object automatically generates four diagnostic residuals plots.

```{r}
par(mfrow=c(2,2)) # Change graphics to a 2x2 graphical matrix
plot(CO2.lm2) # Plot residuals
par(mfrow=c(1,1)) # Return graphics to default
```

In the first plot the residuals appear to be randomly distributed along the fitted values, suggesting that there is a linear relationship between `uptake` and the other variables in our model.  There is some tailing in the Normal Q-Q plot, so the residuals may not follow a perfectly normal distribution.  In the Scale-Location plot the residuals appear to have a relatively horizontal spread so we can assume that there is equal variance.  In the final plot, all of the residuals are within the Cook's distance so we can assume there are no points with high leverage affecting the fit of our model.  So while the our model may not have a perfect fit (which is extremely rare when working with real-world data) the linear regression model does appear to fit well for our data.

From the exploratory plots of our data we may also be interested if there are any interactions between our two categorical factors (`Type` and `Treatment`) and `conc`.  Try fitting and interpreting a linear regression model that has these interactions and compare them to the results from [an ANOVA or ANCOVA model](https://tylerbg.github.io/CDAR/docs/CO2_ANOVA).

*Note: The results from the linear regression model using `lm()` are the same as those from the ANOVA model using `aov()` for the model without interactions.  However, you might notice that the model with the `Type x Treatment` interaction term has different statistical results between the summaries for `lm()` and `aov()`.  This is because of the types of [sum of squares](https://www.r-bloggers.com/2011/03/anova-%E2%80%93-type-iiiiii-ss-explained/) used where the model summary for `aov()` uses type I and for `lm()` uses type III sum of squares.*

\newline
\newline

### Other Resources

[Penn State: Regression Methods](https://online.stat.psu.edu/stat501/lesson/welcome-stat-501)

[STHDA: Multiple Linear Regression in R](http://www.sthda.com/english/articles/40-regression-analysis/168-multiple-linear-regression-in-r/)

[Applied Statistics with R: Multiple Linear Regression](https://daviddalpiaz.github.io/appliedstats/multiple-linear-regression.html)