---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='CO2_ANOVA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
options(width=100)
```

# Analyzing `CO2` with linear regression and ANOVA

To determine which, if any, of the variables have a significant relationship with CO<sub>2</sub> uptake we can use two powerful statistical tools, linear regression and Analysis of Variance (ANOVA).



In our case with two categorical and one continuous variable, our equation will look like:

$$y_{i, \textit{uptake}} = {\beta}_0 + {\beta}_\textit{Treatment} x_{i,\textit{Treatment}} + {\beta}_\textit{Type} x_{i,\textit{Type}} + {\beta}_\textit{conc} x_{i,\textit{conc}} + {\epsilon}_i$$

When using categorical data to fit a linear regression model we would need to [recode the variables](https://stats.idre.ucla.edu/spss/faq/coding-systems-for-categorical-variables-in-regression-analysis/) to numerical values so that they may be entered into the regression equation.  Thankfully, `R` automatically dummy codes factors when they are added to regression model, so we can go ahead and generate and analyze our model using the following code.

```{r}
CO2.lm <- lm(uptake~Type+Treatment+conc, data=CO2)

summary(CO2.lm)
```

Looking at our summary statistics we can see that the linear model has a statistically significant fit (p < 2.2e-16) with a fairly high R<sup>2</sup> (0.6839) and adjusted R<sup>2</sup> (0.6721).  From the p-values for our estimated coefficients (`Pr(>|t|)`), we can see that each of our variables are significant predictors of CO<sub>2</sub> uptake.  Therefore, we can conclude that our three predictor variables, `conc`, `Treatment`, and `Type`, are significant predictors of `uptake`.  We can then take the significant coefficients to rewrite the linear regression equation to:

$${y_{i, \textit{uptake}} = 29.260 + 0.018 x_{i,\textit{Treatment}} - 6.860 x_{i,\textit{Type}} - 12.660 x_{i,\textit{conc}}}$$

To see how `R` coded the categorical variables we can call `$xlevels` from the `lm()` object we just created.

```{r}
CO2.lm$xlevels
```

From the output we can see that for the Type and Treatment variables `Quebec` and `nonchilled` are first followed by `Mississippi` and `chilled`, respectively.  Knowing that R uses dummy coding for factors in regression we can conclude that the first levels are coded as 0 and the second as 1 for the two variables, which we can see further when we use `summary()` to see the results of our regression model.

Notice that the Type and Treatment variables are concatenated with `Mississippi` and `chilled`, respectively.  When Type is `Quebec`, the dummy code is 0 and therefor the value will be 0 in the regression equation ($\beta_3 * x_{i,\textit{Type}} = -12.659 * 0 = 0$).  Conversely, when Type is `Mississippi` which is dummy coded to 1 the value for that variable will equal the estimated coefficient ($\beta_3 * x_{i,\textit{Type}}=-12.659 * 1 = -12.659$).  If we happened to have a third level in Type, the value for that factor would be double the estimated coefficient, triple for a fourth level, and so forth.

```{r}
CO2.lm <- lm(uptake~Type*Treatment+conc, data=CO2)

summary(CO2.lm)
```





```{r}
CO2.aov <- aov(CO2.lm)
```

Although the `conc` variable has 7 distinct levels, molecular concentrations are continuous and it would be sensible to model conc as a continuous variable instead of a categorical one. We can easily do this without making changes to the data by calling `factor()` on the variable within the model formula.

```{r}
CO2.aov <- aov(uptake~Type*Treatment*factor(conc), data=CO2)
```

*Note: Because linear regression and ANOVA are only different in language but not math, you could also use `aov()` on a previously fit `lm()` object.*

Using `summary()`, or alternatively `anova()`, we can print out the ANOVA table for our fitted linear model.

```{r}
summary(CO2.aov)
```

*Note: I have seen others suggest that both `aov()` and `anova()` fit an ANOVA model in `R`, but that is not exactly true.  While `anova()` prints an ANOVA table from a fitted model, only `aov()` actually fits an ANOVA model in base `R`.*



\newline
\newline

### Other Resources

[STHDA: One-Way ANOVA Test in R](http://www.sthda.com/english/wiki/one-way-anova-test-in-r)

[Scribbr: ANOVA in R: A step-by-step guide](https://www.scribbr.com/statistics/anova-in-r/)

[Data Analysis in R: Understanding ANOVA in R](https://bookdown.org/steve_midway/DAR/understanding-anova-in-r.html)

[Datanovia: ANOVA in R](https://www.datanovia.com/en/lessons/anova-in-r/)

[Applied Statistics with R: Analysis of Variance](https://daviddalpiaz.github.io/appliedstats/analysis-of-variance.html)