---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='CO2_EDA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="imgs/CO2_EDA/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# Exploratory Data Analysis of CO2

Before we even start analyzing our data we should get an idea of what is actually in the dataset; what was measured, what are the dependent and independent variables, etc.  We will typically know this before we load our dataset, but for the datasets that are provided in base `R` and other libraries we can simply use `?` followed by the name of the dataset to see the documentation of the dataset.

```{r info}
?CO2
```

We can see that there are five variables within the dataset:

- `Plant`: a unique identifier for each observation,
- `Type`: a categorical variable that gives the origin for each plant,
- `Treatment`: a categorical variable that indicates which temperature treatment the plant was given,
- `conc`: the concentration of CO<sub>2</sub> provided to the plant, and
- `uptake`: the uptake rate of CO<sub>2</sub> by the plant.

We can confirm the variable types and their levels by using `str()` with the dataset.

```{r str}
str(CO2)
```

We see that `Type` and `Treatment` are both categorical variables (factors in `R`) each with 2 levels while `conc` and `uptake` are both continuous variables (numerical in `R`).  `Plant` is also listed as an ordinal factor, however as it is a unique identifier we really do not have any interest in this variable for analysis.

We can better see the distribution of our variables using the `summary()` command.

```{r summary}
summary(CO2)
```

Because both `Type` and `Treatment` are factors `R` prints out the number of times each of the factor levels are present in the dataset (42 times for each level in both factors).  The numerical variables, `conc` and `uptake`, have statistics of their distributions printed out, namely their minimum and maximum values, 1<sup>st</sup>, 2<sup>nd</sup> (median), and 3<sup>rd</sup> quartiles, and their average (mean) value.  We can also use some simple plots in base `R` to graphically observe the relationships between the variables.  Using `plot()` with provide a set of scatterplots that plot each variable against one another.

```{r}
plot(CO2)
```

We can also graphically observe the distribution of each variable with tools such as `hist()` and `boxplot()` that print out a histogram and a boxplot of the variable of interest, respectively.

```{r hist}
par(mfrow=c(1,2))
hist(CO2$uptake)
boxplot(CO2$uptake)
par(mfrow=c(1,1))
```

If we want to determine if a variable follows a normal distribution, `qqnorm()` followed by `qqline()` will print out a normalized Q-Q plot.

```{r}
qqnorm(CO2$uptake)
qqline(CO2$uptake)
```


```{r}
plot(CO2$conc, CO2$uptake)
abline(lm(uptake~conc, CO2))
```


\newline
\newline

### Other Resources

[STHDA: Exploratory Data Analysis](http://www.sthda.com/english/wiki/exploratory-data-analysis)