---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='CO2_ANOVA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
options(width=100)
```

# Analyzing ``CO2`` with linear regression and ANOVA

To determine which, if any, of the variables have a significant relationship with CO<sub>2</sub> uptake we can use two powerful statistical tools, linear regression and Analysis of Variance (ANOVA).

Although the conc variable has 7 distinct levels, molecular concentrations are continuous and it would be sensible to model conc as a continuous variable instead of a categorical one.  

<center><i>y<sub>i</sub> = &beta;<sub>0</sub> + &beta;<sub>1</sub>x<sub>i,1</sub> + &beta;<sub>2</sub>x<sub>i,2</sub> + ... + &beta;<sub>n</sub>x<sub>i,n</sub> + &epsilon;<sub>i</sub></center></i>

- <i>y<sub>i</sub></i> is the response variable
- <i>x<sub>i</sub></i> are our predictor variables
- <i>&beta;</i> are coefficients for the predictor variables and the intercept
- <i>&epsilon;<sub>i</sub></i> is the error term

In our case, our equation will look like:

<center><i>
y<sub>i,uptake</sub> = &beta;<sub>0</sub> + &beta;<sub>conc</sub>x<sub>i,conc</sub> + &beta;<sub>Treatment</sub>x<sub>i,Treatment</sub> + &beta;<sub>Type</sub>x<sub>i,Type</sub> + &epsilon;<sub>i</sub>
</center></i>

When using categorical data to fit a linear regression model we would need to [recode the variables](https://stats.idre.ucla.edu/spss/faq/coding-systems-for-categorical-variables-in-regression-analysis/) to numerical values so that they may be entered into the regression equation.  Thankfully, `R` automatically dummy codes factors when they are added to regression model, so we can go ahead and generate and analyze our model using the following code.

```{r}
CO2.lm <- lm(uptake~Type+Treatment+conc, data=CO2)

summary(CO2.lm)
```

Looking at our summary statistics we can see that the linear model has a statistically significant fit (p < 2.2e-16) with a fairly high R<sup>2</sup> (0.6839) and adjusted R<sup>2</sup> (0.6721).  From the p-values for our estimated coefficients (`Pr(>|t|)`), we can see that each of our variables are significant predictors of CO<sub>2</sub> uptake.  Therefore, we can conclude that our three predictor variables, concentration, Treatment, and Type, are significant predictors of CO<sub>2</sub> uptake.  We can then rewrite our regression equation to:

<center><i>
y<sub>i,uptake</sub> = 29.260 + 0.018x<sub>i,conc</sub> - 6.860x<sub>i,Treatment</sub> - 12.660x<sub>i,Type</sub>
</center></i>

To see how `R` coded the categorical variables we can call `$xlevels` from the `lm()` object we just created.

```{r}
CO2.lm$xlevels
```

From the output we can see that for the Type and Treatment variables "Quebec" and "nonchilled" are first followed by "Mississippi" and "chilled", respectively.  Knowing that R uses dummy coding for factors in regression we can conclude that the first levels are coded as 0 and the second as 1 for the two variables, which we can see further when we use `summary()` to see the results of our regression model.

Notice that the Type and Treatment variables are concatenated with "Mississippi" and "chilled", respectively.  When Type is "Quebec", the dummy code is 0 and therefor the value will be 0 in the regression equation ($\beta_3*x_{i,Type}=-12.659 * 0 = 0$).  Conversely, when Type is "Mississippi" which is dummy coded to 1 the value for that variable will equal the estimated coefficient ($\beta_3*x_{i,Type}=-12.659 * 1 = -12.659$).  If we happened to have a third level in Type, the value for that factor would be double the estimated coefficient, triple for a fourth level, and so forth.






```{r}
CO2.aov <- aov(CO2.lm)
```



```{r}
anova(CO2.aov)
```

\newline
\newline

### Other Resources

[STHDA: One-Way ANOVA Test in R](http://www.sthda.com/english/wiki/one-way-anova-test-in-r)

[Scribbr: ANOVA in R: A step-by-step guide](https://www.scribbr.com/statistics/anova-in-r/)

[Data Analysis in R: Understanding ANOVA in R](https://bookdown.org/steve_midway/DAR/understanding-anova-in-r.html)

[Datanovia: ANOVA in R](https://www.datanovia.com/en/lessons/anova-in-r/)

[Applied Statistics with R: Analysis of Variance](https://daviddalpiaz.github.io/appliedstats/analysis-of-variance.html)

[Penn State: Regression Methods](https://online.stat.psu.edu/stat501/lesson/welcome-stat-501)