---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='one-way-ANOVA.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="images/R/one-way-ANOVA/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# One-way Analysis of Variance (ANOVA)

### [ANOVA hypotheses](https://online.stat.psu.edu/stat500/lesson/10/10.2)

&nbsp;&nbsp;&nbsp;&nbsp;**Null hypothesis:** the means of each group are equal.  
&nbsp;&nbsp;&nbsp;&nbsp;**Alternative hypothesis:** at least one of the group means is not equal to the others.

### [ANOVA assumptions](https://online.stat.psu.edu/stat500/lesson/10/10.2/10.2.1)

- The observations are **independent**.
- Each factor is approximately **normally distributed**.
- The distributions the factors have **equal variances**.

### Entering the data in R

&nbsp;&nbsp;&nbsp;&nbsp;For this tutorial we will be using a simple dataset from the [PSU STAT 500 Applied Statistics course](https://online.stat.psu.edu/stat500/lesson/10/10.1). Typically, we would load our data into R from a previously generated file, however for this example we will manually input the data into a data frame with the following code.

```{r}
lab_precise <- data.frame(tar_mg = c(10.21, 10.25, 10.24, 9.80, 9.77, 9.73,
                                     11.32, 11.20, 11.40, 10.50, 10.68, 10.90,
                                     11.60, 11.90, 11.80, 12.30, 12.20, 12.20),
                          Brand = c(rep("Brand A", 6),
                                    rep("Brand B", 6),
                                    rep("Brand C", 6)))
```

Using the `data.frame()` function we create a data frame with two columns, one labeled *tar_mg* with the tar content from each sample and the second *Brand* which indicates which cigarette brand each measurement came from.  Note that the data frame is currently in long format which is preferred in many R functions compared to the wide format as shown in the [PSU STAT 500 example](https://online.stat.psu.edu/stat500/lesson/10/10.1).

### Fitting an one-way ANOVA in R

&nbsp;&nbsp;&nbsp;&nbsp;Now that we have our data, we can fit an ANOVA model using the `aov()` function in R with the following syntax:

- A formula for our model with the response on the left and the factor names to the right of a `~`
- Specification of the dataset to be used

```{r}
aov.fit <- aov(tar_mg ~ Brand, data = lab_precise)
```

### Interpreting the results

&nbsp;&nbsp;&nbsp;&nbsp;To summarize the results of the ANOVA model we can use the `summary()` or alternatively `anova()` function with the fitted model as the only argument.

```{r}
summary(aov.fit)
```

&nbsp;&nbsp;&nbsp;&nbsp;From the summary of the ANOVA model we see that our factor, *Brand*, has a p-value well below our threshold of 0.05. This tells us that the mean tar content for at least one of the brands of cigarettes is different from the others. To determine which of the means are different from one another we follow up the ANOVA with a post-hoc test, in this case we are using a Tukey's honestly significant difference test (Tukey HSD) with the `TukeyHSD()` function.

```{r}
TukeyHSD(aov.fit)
```

&nbsp;&nbsp;&nbsp;&nbsp;From the Tukey HSD post-hoc test we see that the p-values for each comparison are below 0.05, meaning that each brand of cigarettes is significantly different from one another.

### Full code block

```{r eval=FALSE}
# Enter the data
lab_precise <- data.frame(tar_mg = c(10.21, 10.25, 10.24, 9.80, 9.77, 9.73,
                                     11.32, 11.20, 11.40, 10.50, 10.68, 10.90,
                                     11.60, 11.90, 11.80, 12.30, 12.20, 12.20),
                          Brand = c(rep("Brand A", 6),
                                    rep("Brand B", 6),
                                    rep("Brand C", 6)))

# Fit a one-way ANOVA model
aov.fit <- aov(tar_mg ~ Brand, data = lab_precise)

# Summarize the results of the ANOVA fit
summary(aov.fit)

# Perform and summarize a Tukey HSD post-hoc test
TukeyHSD(aov.fit)
```