---
title: "Outliers"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'Outliers.md'))})
date: "Published: 4/6/2020"
output: github_document
---

$G = \frac{Y_{max}-\bar{Y}}{sd}$ or $G = \frac{\bar{Y}-Y_{min}}{sd}$

To use this test, we first calculat $G$, our test statistic, from the positive difference of the mean, $\bar{Y}$ with the maximum $Y_{max}$ or minimum $Y_{min}$ value (depending on if our outlier is above or below the average) then divide by the standard deviation, $sd$.  Then we compare our test statistic to a [critical value we calculate for the test](https://www.statisticshowto.com/grubbs-test/). We can use the following code to calculate both $G$ and our critical value then print them out to compare.

```{r}
G <- (max(beer$beer_abv, na.rm=TRUE) - mean(beer$beer_abv, na.rm=TRUE)) / sd(beer$beer_abv, na.rm=TRUE)
N <- length(beer$beer_abv[!is.na(beer$beer_abv)])
G.t <- qt(0.05/N, df=N-2)
G.crit <- (N-1)/sqrt(N) * sqrt(G.t^2 / (N-2+G.t^2))

G
G.crit
```

$G$ is much greater than our critical value, therefore we could conclude that 89 is an outlier.  However, why write lengthy code like we did above when packages exist to do the test for us?  Instead, let us use the `outliers` package to perform a Grubb's test.

*Note: If you have not already installed a package you will first need to do so with `install.packages()`*

```{r}
library(outliers)
grubbs.test(beer$beer_abv)
```

This saves us a lot of coding and gives us the same answer (but this time in p-value form).  With a p-value of 2.2^{-16} we conclude that 89 is a statistically significant outlier.  But what if we had two or more outliers?  