---
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
                      encoding=encoding,
                      output_file='ToothGrowth_ANOVA') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="imgs/ToothGrowth_ANOVA/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# EDA with `ToothGrowth`

$${y_\textit{ijk} = {\mu}_{..} + {\alpha}_i + {\beta}_j + ({\alpha}{\beta})_\textit{ij} + {\epsilon}_\textit{ijk}}$$
Where:
- $y$ is the response with $i$ and $j$ factor levels and $k$ observations,
- $\mu$ is the grand mean,
- $\alpha$ and $\beta$ are factors with $i$ and $j$ levels respectively,
- $(\alpha\beta)$ is the interaction term for $i$ and $j$ levels, and
- $\epsilon$ is the error term with $i$ and $j$ factor levels and $k$ observations.

$${\textit{len}_\textit{ijk} = {\mu}_{..} + \textit{supp}_i + \textit{dose}_j + (\textit{supp}*\textit{dose})_\textit{ij} + {\epsilon}_\textit{ijk}}$$

```{r}
tg.aov <- aov(len~supp*factor(dose), data=ToothGrowth)

summary(tg.aov)
```

First, we assess the interaction term `supp:factor(dose)` which is statistically significant.

### Other Resources