---
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file='two-proportions.md') })
output: github_document
editor_options:
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="imgs/two-proportions/",
                      fig.width=6, fig.height=4,
                      fig.show='hold', fig.align='center')
options(width=100)
```

# Comparing Two Proportions

#### Hypotheses

&nbsp;&nbsp;&nbsp;&nbsp;*Null hypothesis:* the two proportions are equal.  
&nbsp;&nbsp;&nbsp;&nbsp;*Alternative hypothesis:* the two proportions are not equal.

#### Assumptions


#### Running the test

For this R tutorial we will use the example from [Penn State's Introduction to Mathematical Statistics course](https://online.stat.psu.edu/stat415/lesson/9/9.4).  Assume that 800 adult Americans were polled on the question: "Should the federal tax on cigarettes be raised to pay for health care reform?".

```{r}
n1 <- 605
n2 <- 195
y1 <- 351
y2 <- 41

prop.test(x = c(y1, y2), n = c(n1, n2))
```

